### 2018-06-08 doc/main.tex
Added source for documentation

\documentclass[a4paper,parskip=never,UKenglish]{scrartcl}
\input{preamble}
\title{Rusty Variation}
\subtitle{Affine Session Types for Rust}
\author{%
  Wen Kokke\protect\\
  Supervised by Philip Wadler and Ian Stark\protect\\
  Graduate student (8450698)}
\affil{
  LFCS, University of Edinburgh\protect\\
  Informatics Forum, 10 Crichton St, Edinburgh EH8 9AB, UK\protect\\
  \email{wen.kokke@ed.ac.uk}}
\date{June 16, 2018}
\begin{document}
\maketitle
\begin{abstract}
  Rusty Variation (RV) is a library for session-typed communication in Rust, which offers strong compile-time correctness guarantees. Programs written using RV are guaranteed to adhere to a specified protocol, and are guaranteed to be free from dead- and livelocks. RV is an implementation of Exceptional GV, a session-typed functional language with exceptions and session cancellation.
\end{abstract}

\section{Problem and Motivation}

\section{Background and Related Work}
\subsection{Session Types for Rust}
\subsection{Exceptional GV}

\section{Approach and Uniqueness}

\section{Results and Contributions}
\end{document}

### 2018-06-08 doc/main.tex
Added Rust code example and listings

@@
 \input{preamble}
+\addbibresource{main.bib}
 \title{Rusty Variation}
@@
   \email{wen.kokke@ed.ac.uk}}
-\date{June 16, 2018}
+\date{}
 \begin{document}
@@
 \subsection{Session Types for Rust}
+\textcite{jespersen2015}
+
 \subsection{Exceptional GV}
+\textcite{fowler2018}
 
 \section{Approach and Uniqueness}
+\begin{lstlisting}[language=Rust, style=boxed]
+  use std::marker;
+  use rusty_variation::*;
+
+  type NegServer  = Receive<i32, Send<i32, End>>;
+  type AddServer  = Receive<i32, Receive<i32, Send<i32, End>>>;
+  type CalcServer = Receive<CalcOp, End>;
+  enum CalcOp { Neg(NegServer), Add(AddServer) }
+
+  #[test]
+  fn calc_server_works() {  
+    assert!(|| -> Result<i32, Box<Error>> {
+
+      // Fork a calculator server.
+      let s = fork!(move |s: CalcServer| {
+        offer!(s, {
+          CalcOp::Neg(s) => {
+            let (x, s) = receive(s)?;
+            let s = send(-x, s)?;
+            close(s)
+          },
+          CalcOp::Add(s) => {
+            let (x, s) = receive(s)?;
+            let (y, s) = receive(s)?;
+            let s = send(x + y, s)?;
+            close(s)
+          },
+        })
+      });
+
+      // Send it the numbers. 
+      let s = select!(CalcOp::Add, s)?;
+      let s = send(4, s)?;
+      let s = send(5, s)?;
+
+      // Receive the answer and close the channel.
+      let (z, s) = receive(s)?;
+      close(s)?;
+
+      // Check the answer.
+      assert_eq!(z, 9);
+
+      Ok(())
+
+  }().is_ok());  
+\end{lstlisting}
 
 \section{Results and Contributions}
+
+\clearpage
+\printbibliography
 \end{document}

### 2018-06-14 doc/main.tex
Draft version of SRC submission

--- 2018-06-08 doc/main.tex
@@
-\documentclass[a4paper,parskip=never,UKenglish]{scrartcl}
+\documentclass[sigconf,UKEnglish,natbib=false]{acmart}
 \input{preamble}
+
+\begin{document}
 \title{Rusty Variation}
-\subtitle{Affine Session Types for Rust}
-\author{%
-  Wen Kokke\protect\\
-  Supervised by Philip Wadler and Ian Stark\protect\\
-  Graduate student (8450698)}
-\affil{
-  LFCS, University of Edinburgh\protect\\
-  Informatics Forum, 10 Crichton St, Edinburgh EH8 9AB, UK\protect\\
-  \email{wen.kokke@ed.ac.uk}}
-\date{June 16, 2018}
-\begin{document}
-\maketitle
+\subtitle{Deadlock-free Sessions with Failure in Rust}
+\author{Wen Kokke}
+\authornote{%
+  Graduate student (8450698), supervised by Philip Wadler and Ian Stark.}
+\affiliation{%
+  LFCS, University of Edinburgh, 10 Crichton St,
+  EH8 9AB, Edinburgh, United Kingdom}
+\email{wen.kokke@ed.ac.uk}
+
 \begin{abstract}
-  Rusty Variation (RV) is a library for session-typed communication in Rust, which offers strong compile-time correctness guarantees.
-Programs written using RV are guaranteed to adhere to a specified protocol, and are guaranteed to be free from dead- and livelocks.
-RV is an implementation of Exceptional GV, a session-typed functional language with exceptions and session cancellation.
+  Rusty Variation (RV) is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees.
+Programs written using RV are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
 \end{abstract}
 
-\section{Problem and Motivation}
+\maketitle
 
-\section{Background and Related Work}
-\subsection{Session Types for Rust}
-\subsection{Exceptional GV}
+\section*{Problem \& Motivation}
+In concurrent programming, processes commonly exchange information by sending messages on shared channels.
+However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.
 
-\section{Approach and Uniqueness}
+Session types are a substructural typing discipline, introduced by \textcite{honda1993}, which describe communication protocols as \emph{channel types}, capturing both the type and the order of messages.
+Session type systems, then, ensure that programs correctly implement specified protocols.
 
-\section{Results and Contributions}
+Rust is an immensely popular systems programming language.\footnote{https://insights.stackoverflow.com/survey/2018/} It aims to combine efficiency with abstraction and thread and memory safety.
+However, out of the box, it does not offer a way to specify protocols and verify communication.
+
+In this paper, we present Rusty Variation, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and races.
+We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}.\footnote{\url{https://doc.rust-lang.org/nightly/book/}}
+
+\section*{Background \& Related Work}
+\Textcite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by \textcite{gay2009}.
+Their implementation guarantees that programs correctly implement specified communication protocols with some caveats.
+LAST depends upon linear types, meaning that sessions have to be used exactly once.
+Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped.
+The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}.
+In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \texttt{\#[must\_use]} annotation.
+However, in general, this is not something the compiler can guarantee.
+Furthermore, LAST ignores the possibility of communication failures, such as network errors, which cannot be detected at compile-time, yet can cause a channel to be dropped.
+
+We embrace the affine, and base our work on Exceptional GV by \textcite[EGV]{fowler2018}.
+Exceptional GV is a descendant of Good Variation~\parencite[GV]{wadler2012}, which itself is a descendant of LAST.
+Crucially, EGV has an affine type system, and allows for the explicit cancellation of sessions.
+Furthermore, EGV offers a stronger metatheory than LAST.
+In addition to session fidelity, EGV also guarantees global progress, freedom from deadlocks and livelocks, confluence, and termination.
+That is to say, there are no locks or race conditions, and all programs eventually halt.
+For a discussion of EGV and its metatheory, we refer the reader to the work by \textcite{fowler2018}.
+
+\section*{Contribution: Rusty Variation}
+Our contribution, Rusty Variation, is an implementation of EGV in Rust. We claim that it is a faithful implementation, which preserves much of EGVs metatheory. However, we depart from it in two ways:
+
+First, EGV relies upon exceptions and handlers.
+Rust, however, has rejected exceptions in favour of monadic errors i.e., wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.
+
+Second, the core language of EGV does not include general recursion.
+Rust does.
+In the presence of non-termination, we cannot guarantee session fidelity, livelock freedom, or---well---termination.
+The work by \textcite{jespersen2015} also suffers from this shortcoming.
+We can demonstrate this by writing the program which loops forever, and then sends a message.
+The Rust type checker does not see a problem with this, but obviously the message will never be sent.
+However, we claim that if the program terminates, it respected the communication protocol.
+
+\subsection*{Sessions and duality}
+Any type built using the primitive types below is a valid session:
+\begin{lstlisting}
+pub struct End;
+pub struct Send<T, S: Session>{ ... }
+pub struct Recv<T, S: Session>{ ... }
+\end{lstlisting}
+Session \emph{types} are public, but their members are not.
+Therefore, user cannot construct their own misbehaving sessions.
+The continuation of a session must itself be a session.
+This is enforced by the \lstinline{Session} constraint.
+In addition to this ``kinding'' of session types, the \lstinline{Session} trait also implements duality and session generation:
+\begin{lstlisting}
+pub trait Session {
+  type Dual: Session<Dual=Self>;
+  fn new() -> (Self, Self::Dual);
+}
+\end{lstlisting}
+Duality works as expected: \lstinline{Send<T, S>} is dual to \lstinline{Recv<T, S::Dual>}---and vice versa---and \lstinline{End} is self-dual.
+The constraint \lstinline{Session<Dual=Self>} enforces that the dual of a
+session is a session, and that \lstinline{S::Dual::Dual} is equal to \lstinline{S} (involutivity).
+
+The \lstinline{new} function generates a new session and returns two \emph{dual} channel endpoints.
+Unlike the rest of the functions discussed in this section, \lstinline{new} is \emph{not} part of the public interface.
+
+Using these types, we can define the types of servers which offer negation and addition as follows, where \lstinline{i32} is the type of 32-bit integers:
+\begin{lstlisting}
+type NegSrv = Recv<i32, Send<i32, End>>;
+type AddSrv = Recv<i32, Recv<i32, Send<i32, End>>>;
+\end{lstlisting}
+The client types are obtained by duality.
+
+\subsection*{Send, receive, and fork!}
+The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. They have the following types:
+\begin{lstlisting}
+pub fn send<T, S: Session>(x: T, s: Send<T, S>)
+       -> Result<S, Box<Error>>;
+pub fn recv<T, S: Session>(s: Recv<T, S)
+       -> Result<(T, S), Box<Error>>;
+\end{lstlisting}
+The \lstinline{fork!} macro creates a new session, and spawns a new thread.
+The new thread runs the first argument of \lstinline{fork!}---a code block---which receives one endpoint, and \lstinline{fork!} returns the dual endpoint.
+While \lstinline{fork!} is implemented as a macro, it can be considered as a function with the following type:
+\begin{lstlisting}
+pub fn fork!<S: Session>(p: P) -> S::Dual
+      where
+        P: FnOnce(S) -> Result<(), Box<Error>>;
+\end{lstlisting}
+Using these primitives, we can write the following program. This program forks off a process which sends a ``ping'' along the session, then waits to receive that ping and returns:
+\begin{lstlisting}
+let s = fork!(move |s: Send<(), End>| { send((), s) });
+let ((), End) = recv(s)?;
+Ok(())
+\end{lstlisting}
+The \lstinline{let x = M?;} construct is Rust's ``monadic bind'' notation for programs which may return errors.
+If \lstinline{recv(s)} fails, the \lstinline{?}-operator short-circuits, skips the rest of the statements, and returns the error.
+
+\subsection*{Cancellation}
+The \lstinline{cancel} function drops a value and cancels all sessions \emph{and} their duals referenced in that value.
+This cancellation propagates: any process attempting to write or write to a cancelled channel will return fail and cancel all \emph{their} open channels.
+This ensures that no process is left waiting for a message that will never come.
+\begin{lstlisting}
+pub fn cancel<T>(x: T) -> Result<(), Box<Error>>;
+\end{lstlisting}
+Using \lstinline{cancel}, we can implement the following program.
+This program forks off a process which it \emph{expects} to send a ``ping''.
+However, instead the forked process cancels the session.
+Therefore, the call to \lstinline{recv} fails, the \lstinline{?}-operator short-circuits, and the whole process returns an error:
+\begin{lstlisting}
+let s = fork!(move |s: Send<(), End>| { cancel(s) });
+let (z, End) = recv(s)?;
+Ok(())
+\end{lstlisting}
+The cancellation mechanism is implemented through destructors. In fact, in the current implementation it is inherited from the implementation of channels in \lstinline{std::sync::mpsc}.
+
+The \lstinline{Send} and \lstinline{Recv} structures are annotated with \lstinline{#[must_use]}, to encourage the user to complete the protocol or cancel it explicitly.
+However, should a session be dropped, this is interpreted as \emph{implicit} cancellation.
+This ensures deadlock freedom, even in Rust's affine type system.
+
+\subsection*{Branching with offer! and select!}
+The \lstinline{offer_either}, \lstinline{select_left}, and \lstinline{select_right} functions let processes offer or make a \emph{binary} choice.
+They implement this by receiving or sending a value of the sum type \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.
+For instance, we could write the type of a ``calculator'' server, offering negation and addition, as \lstinline{Recv<Either<NegSrv, AddSrv>, End>}.
+
+While these functions suffice to encode choice, we also define two macros, \lstinline{offer!} and \lstinline{select!}, which generalise this functionality to \emph{all enums}---as long as each value wraps a session.
+This allows processes to offer and select choices between many \emph{labelled} branches, and allows us to define the type for our calculator server as:
+\begin{lstlisting}
+type CalcSrv = Recv<CalcOp, End>;
+enum CalcOp {Neg(NegSrv), Add(AddSrv)}
+\end{lstlisting}
+Using these macros, we can write the following program.
+This program forks off a calculator server, which offers the choice between negation and addition, selects addition, sends some numbers, and then returns the result:
+\begin{lstlisting}
+let s = fork!(move |s: CalcSrv| {
+  offer!(s, {
+    CalcOp::Neg(s) => {
+      let (x, s) = recv(s)?;
+      send(-x, s)
+    },
+    CalcOp::Add(s) => {
+      let (x, s) = recv(s)?;
+      let (y, s) = recv(s)?;
+      send(x + y, s)
+    },
+  })
+});
+let s = select!(CalcOp::Add, s)?;
+let s = send(4, s)?;
+let s = send(5, s)?;
+let (z, End) = recv(s)?;
+Ok(z)
+\end{lstlisting}
+
+\subsection*{Limitations}
+There are two issues which can make code written using Rusty Variation deadlock, and surprisingly, \lstinline{unsafe} is not one of them:
+
+The biggest issue lies in the fact that Rust's semantics do not guarantee that destructors are actually run.
+For instance, it is always possible to ``forget'' a value, using \lstinline{std::mem::forget}, without running its destructors.
+The following simple program deadlocks:
+\begin{lstlisting}
+let s = fork!(move |s: Send<(), End>| { mem::forget(s) });
+let ((), End) = recv(s)?;
+Ok(())
+\end{lstlisting}
+
+Second, it is not possible---at the moment---to export a trait without exporting all its members.
+Therefore, any user can use the \lstinline{new} function, and doing so, it is trivial to write a program which deadlocks:
+\begin{lstlisting}
+let (s1, r1) = Send<Void, End>::new();
+let r2 = fork!(move |s2: Send<Void, End>| {
+  let (v, End) = recv(r1)?;
+  send(v, s2)
+});
+let (v, End) = recv(r2)?;
+send(v, s1)
+\end{lstlisting}
+
+Last, the implementation Rusty Variation is inefficient.
+We currently use the asynchronous channels from \lstinline{std::sync::mpsc}.
+These are ideal, as they already implement the correct cancellation behaviour.
+Unfortunately, they can send values of only a single type.
+Therefore, in the implementation of \lstinline{send}, we create a new session for the continuation, and send that along with the value.
+This means we create a ``one-shot'' channel for every call to \lstinline{send}.
+
+\clearpage
+\printbibliography
 \end{document}

### 2018-06-14 src/lib.rs

@@ -470,6 +470,19 @@ mod tests {
 
         assert!(other_thread.join().is_ok());
     }
+
+    fn do_not_run_loops_exist() {
+
+        let s = fork!(move |s: Receive<i32, End>| {
+            let (x, s) = receive(s)?;
+            close(s)
+        });
+
+        loop {
+            // Don't need to do anything, just loop!
+        }
+        assert!(send(0, s).is_ok());
+    }
 }
 
 // */

### 2018-06-14 doc/main.tex
Wrote deadlocking examples into the library; refined draft

@@
-\documentclass[sigconf,UKEnglish,natbib=false]{acmart}
+\documentclass[sigconf,natbib=false]{acmart}
 \input{preamble}
@@
 The \lstinline{new} function generates a new session and returns two \emph{dual} channel endpoints.
-Unlike the rest of the functions discussed in this section, \lstinline{new} is \emph{not} part of the public interface.
+Unlike the rest of the functions discussed in this section, \lstinline{new} is \emph{not} part of the public interface---see \cref{sec:limitations}.
 
@@
 \begin{lstlisting}
-let s = fork!(move |s: CalcSrv| {
-  offer!(s, {
+let s = fork!(move |s: CalcSrv| { offer!(s, {
     CalcOp::Neg(s) => {
@@
       send(x + y, s)
-    },
-  })
-});
+    },})});
 let s = select!(CalcOp::Add, s)?;
@@
 
-\subsection*{Limitations}
-There are two issues which can make code written using Rusty Variation deadlock, and surprisingly, \lstinline{unsafe} is not one of them:
+\subsection*{Implementation}
+We use the asynchronous channels from \lstinline{std::sync::mpsc}.
+These are ideal, as they already implement the correct cancellation behaviour.
+However, they can only transfer values of a single type.
+Therefore, we treat them as linear channels---channels which can only be used once---and encode our communication following \textcite{dardha2017}.
+Each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value.
+Fortunately, the channels in \lstinline{std::sync::mpsc} are highly optimised for one-shot usage.
+
+\subsection*{What If I Want to Be Evil?}\label{sec:limitations}
+There are three ways in which you can make code written using Rusty Variation deadlock.
+The first is by looping, which we discussed earlier.
+The other two are both explicit, and surprisingly, \lstinline{unsafe} is not one of them:
 
@@
 
-Last, the implementation Rusty Variation is inefficient.
-We currently use the asynchronous channels from \lstinline{std::sync::mpsc}.
-These are ideal, as they already implement the correct cancellation behaviour.
-Unfortunately, they can send values of only a single type.
-Therefore, in the implementation of \lstinline{send}, we create a new session for the continuation, and send that along with the value.
-This means we create a ``one-shot'' channel for every call to \lstinline{send}.
-
 \clearpage

### 2018-06-14 src/lib.rs

@@ -28,6 +28,7 @@ pub struct Receive<T, S: Session> {
 pub trait Session: marker::Sized + marker::Send {
     type Dual: Session<Dual=Self>;
 
+    #[doc(hidden)]
     fn new() -> (Self, Self::Dual);
 }
 
@@ -471,17 +472,53 @@ mod tests {
         assert!(other_thread.join().is_ok());
     }
 
-    fn do_not_run_loops_exist() {
+    #[allow(dead_code)]
+    fn deadlock_loop() {
 
-        let s = fork!(move |s: Receive<i32, End>| {
-            let (x, s) = receive(s)?;
-            close(s)
+        let s = fork!(move |s: Send<(), End>| {
+            let trick: bool = false;
+            loop {
+                // Don't need to do anything, just loop!
+                if trick { break; }
+            }
+            send((), s)
         });
 
-        loop {
-            // Don't need to do anything, just loop!
-        }
-        assert!(send(0, s).is_ok());
+        || -> Result<(), Box<Error>> {
+            let ((), End) = receive(s)?;
+            Ok(())
+        }().unwrap();
+    }
+
+    #[allow(dead_code)]
+    fn deadlock_forget() {
+
+        let s = fork!(move |s: Send<(), End>| {
+            mem::forget(s);
+            Ok(())
+        });
+
+        || -> Result<(), Box<Error>> {
+            let ((), End) = receive(s)?;
+            Ok(())
+        }().unwrap();
+    }
+
+    #[allow(dead_code)]
+    fn deadlock_new() {
+
+        let (s1, r1) = <Send<(), End>>::new();
+        let r2 = fork!(move |s2: Send<(), End>| {
+            let (x, End) = receive(r1)?;
+            let End = send(x, s2)?;
+            Ok(())
+        });
+
+        || -> Result<(), Box<Error>> {
+            let (x, End) = receive(r2)?;
+            let End = send(x, s1)?;
+            Ok(())
+        }().unwrap();
     }
 }
 

### 2018-06-14 src/lib.rs
Minor changes

@@ -476,12 +476,12 @@ mod tests {
     fn deadlock_loop() {
 
         let s = fork!(move |s: Send<(), End>| {
-            let trick: bool = false;
             loop {
-                // Don't need to do anything, just loop!
-                if trick { break; }
+                // Let's trick the reachability checker
+                if false { break; }
             }
-            send((), s)
+            let End = send((), s)?;
+            Ok(())
         });
 
         || -> Result<(), Box<Error>> {

### 2018-06-15 doc/main.tex
Final version of SRC submission

@@
 
-\section*{Problem \& Motivation}
+\section{Problem \& Motivation}
 In concurrent programming, processes commonly exchange information by sending messages on shared channels.
@@
 
-\section*{Background \& Related Work}
+\section{Background \& Related Work}
 \Textcite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by \textcite{gay2009}.
@@
 The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}.
-In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \texttt{\#[must\_use]} annotation.
+In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \lstinline{#[must_use]} annotation.
 However, in general, this is not something the compiler can guarantee.
@@
 
-\section*{Contribution: Rusty Variation}
+\section{Contribution: Rusty Variation}
 Our contribution, Rusty Variation, is an implementation of EGV in Rust. We claim that it is a faithful implementation, which preserves much of EGVs metatheory. However, we depart from it in two ways:
@@
 
-\subsection*{Sessions and duality}
+\subsection{Sessions and duality}
 Any type built using the primitive types below is a valid session:
@@
 
-\subsection*{Send, receive, and fork!}
+\subsection{Send, receive, and fork!}
 The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. They have the following types:
@@
 
-\subsection*{Cancellation}
+\subsection{Cancellation}
 The \lstinline{cancel} function drops a value and cancels all sessions \emph{and} their duals referenced in that value.
@@
 
-\subsection*{Branching with offer! and select!}
+\subsection{Branching with offer! and select!}
 The \lstinline{offer_either}, \lstinline{select_left}, and \lstinline{select_right} functions let processes offer or make a \emph{binary} choice.
@@
 \begin{lstlisting}
-let s = fork!(move |s: CalcSrv| { offer!(s, {
+let s = fork!(move |s: CalcSrv| {
+  offer!(s, {
     CalcOp::Neg(s) => {
@@
       send(x + y, s)
-    },})});
+    },
+  })});
 let s = select!(CalcOp::Add, s)?;
@@
 
-\subsection*{Implementation}
+\subsection{Implementation}
 We use the asynchronous channels from \lstinline{std::sync::mpsc}.
 These are ideal, as they already implement the correct cancellation behaviour.
-However, they can only transfer values of a single type.
-Therefore, we treat them as linear channels---channels which can only be used once---and encode our communication following \textcite{dardha2017}.
-Each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value.
-Fortunately, the channels in \lstinline{std::sync::mpsc} are highly optimised for one-shot usage.
+Channels are typed, and can only transfer values of a single type.
+Therefore, we use one-shot channels, and encode sessions following \textcite{dardha2017,scalas2016,padovani2017}: each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value, and each \lstinline{recv} receives a value together with a channel on which to continue the session.
+Fortunately, the channels in \lstinline{std::sync::mpsc} are highly optimised for one-shot usage.\footnote{\url{https://doc.rust-lang.org/src/std/sync/mpsc/mod.rs.html\#147-149}}
 
-\subsection*{What If I Want to Be Evil?}\label{sec:limitations}
+\subsection{What If I Want to Be Evil?}\label{sec:limitations}
 There are three ways in which you can make code written using Rusty Variation deadlock.
 The first is by looping, which we discussed earlier.
-The other two are both explicit, and surprisingly, \lstinline{unsafe} is not one of them:
+The other two are explicit, and surprisingly, \lstinline{unsafe} is not one of them:
 

### 2018-06-15 doc/main.tex
Bumped release to v1.0.0
Inserted link to GitHub

@@
 \authornote{%
-  Graduate student (8450698), supervised by Philip Wadler and Ian Stark.}
+  Graduate student (9727050), supervised by Philip Wadler and Ian Stark.}
 \affiliation{%
@@
 
-In this paper, we present Rusty Variation, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and races.
+In this paper, we present Rusty Variation\footnote{\url{https://github.com/wenkokke/rusty-variation}}, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and races.
 We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}.\footnote{\url{https://doc.rust-lang.org/nightly/book/}}

### 2018-08-29 src/lib.rs
Renamed *receive->recv, updated example in README.md

@@ -21,7 +21,7 @@ pub struct Send<T, S: Session> {
 
 #[must_use]
 #[derive(Debug)]
-pub struct Receive<T, S: Session> {
+pub struct Recv<T, S: Session> {
     channel: Receiver<(T, S)>,
 }
 
@@ -41,15 +41,15 @@ impl Session for End {
 }
 
 impl<T: marker::Send, S: Session> Session for Send<T, S> {
-    type Dual = Receive<T, S::Dual>;
+    type Dual = Recv<T, S::Dual>;
 
     fn new() -> (Self, Self::Dual) {
         let (sender, receiver) = mpsc::channel::<(T, S::Dual)>();
-        return (Send { channel: sender }, Receive { channel: receiver });
+        return (Send { channel: sender }, Recv { channel: receiver });
     }
 }
 
-impl<T: marker::Send, S: Session> Session for Receive<T, S> {
+impl<T: marker::Send, S: Session> Session for Recv<T, S> {
     type Dual = Send<T, S::Dual>;
 
     fn new() -> (Self, Self::Dual) {
@@ -68,7 +68,7 @@ pub fn send<'a, T: marker::Send + 'a, S: Session + 'a>(x: T, s: Send<T, S>)
     Ok(here)
 }
 
-pub fn receive<'a, T: marker::Send, S: Session>(s: Receive<T, S>)
+pub fn recv<'a, T: marker::Send, S: Session>(s: Recv<T, S>)
                                                 -> Result<(T, S), Box<Error + 'a>> {
     let (v, s) = s.channel.recv()?;
     Ok((v, s))
@@ -131,7 +131,7 @@ macro_rules! fork {
     };
 }
 
-pub type Offer<S1, S2> = Receive<Either<S1, S2>, End>;
+pub type Offer<S1, S2> = Recv<Either<S1, S2>, End>;
 pub type Select<S1, S2> = Send<Either<<S1 as Session>::Dual, <S2 as Session>::Dual>, End>;
 
 pub fn offer_either<'a, S1: Session, S2: Session, F, G, R>(s: Offer<S1, S2>, f: F, g: G)
@@ -140,7 +140,7 @@ where
     F: FnOnce(S1) -> Result<R, Box<Error + 'a>>,
     G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
 {
-    let (e, End) = receive(s)?;
+    let (e, End) = recv(s)?;
     e.either(f, g)
 }
 
[…]

### 2019-03-15 egv/src/Language/EGV.hs
One step closer to tracer implementation

@@ -1,14 +1,20 @@
+{-# LANGUAGE ViewPatterns #-}
+{-# LANGUAGE PatternGuards #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE RecordWildCards #-}
 {-# LANGUAGE TemplateHaskell #-}
 {-# LANGUAGE EmptyDataDeriving #-}
 {-# LANGUAGE DeriveDataTypeable #-}
 module Language.EGV where
 
-import Control.Applicative (Alternative(empty))
+import Control.Applicative (Alternative(empty,(<|>)))
 import Control.Enumerable (Sized(aconcat), Enumerable(enumerate), share, global)
-import Data.IntMap (IntMap)
+import Data.List (find)
+import Data.IntMap (IntMap,(!))
 import qualified Data.IntMap as IM
 import Data.Typeable (Typeable)
-import Test.Feat (Enumerate, c0, c1, c2, deriveEnumerable)
+import Test.Feat (Enumerate, c0, c1, c2, c3, c4, c5)
 
 
 -- * Types
@@ -53,8 +59,9 @@ instance Enumerable n => Enumerable (S n) where
   enumerate = share $ aconcat [ c0 FZ , c1 FS ]
 
 
-
 -- * Terms
+
+type Name = Int
 
 data Term n
   = Var n
@@ -74,11 +81,16 @@ data Term n
   | Cancel (Term n) Sess
   | Try (Term n) (Term (S n)) (Term n) Type
   | Raise
+  | Chan Name -- ^ Runtime Syntax
   deriving (Typeable, Eq, Show)
 
 
-deriveEnumerable ''Term
-
+instance Enumerable n => Enumerable (Term n) where
+  enumerate = share $ aconcat
+    [ c1 Var  , c1 Lam     , c3 App  , c0 TT    , c2 LetTT
+    , c2 Pair , c4 LetPair , c1 Inl  , c1 Inr   , c5 Case
+    , c1 Fork , c3 Send    , c1 Recv , c1 Close , c2 Cancel
+    , c4 Try  , c0 Raise ]
 
 closed :: Enumerate (Term Z)
 closed = global
@@ -192,22 +204,7 @@ checkClosed :: Type -> Term Z -> Bool
 checkClosed = check IM.empty
 
 
--- * Configurations
-
-data Flag
-  = Main
-  | Child
-
[…]

### 2019-03-17 tests/session_types_tests.rs
Added segfault tests for session-types library.

@@ -1,0 +1,49 @@
+#![feature(test)]
+
+extern crate session_types;
+extern crate test;
+extern crate signal_hook;
+
+#[cfg(test)]
+#[allow(dead_code)]
+mod session_types_tests {
+
+    use std::boxed::Box;
+    use std::error::Error;
+    use std::thread;
+    use session_types::*;
+    use signal_hook::{iterator::Signals, SIGSEGV};
+
+    fn client(c: Chan<(), Send<(), Eps>>) {
+        c.send(()).close();
+    }
+
+    fn server(c: Chan<(), Recv<(), Eps>>) {
+        let (c, ()) = c.recv();
+        c.close();
+    }
+
+    fn drop_client(_c: Chan<(), Send<(), Eps>>) {}
+
+    fn drop_server(_c: Chan<(), Recv<(), Eps>>) {}
+
+    #[test]
+    fn server_client_works() {
+        connect(server, client);
+    }
+
+    // #[test]
+    fn client_incomplete_panics() {
+        connect(server, drop_client);
+    }
+
+    #[test]
+    fn server_incomplete_segfaults() {
+        connect(drop_server, client);
+    }
+
+    // #[test]
+    fn server_client_incomplete_panics() {
+        connect(drop_server, drop_client);
+    }
+}

### 2019-03-21 tests/session_types_tests.rs
Disabled faulty test

@@ -2,7 +2,6 @@
 
 extern crate session_types;
 extern crate test;
-extern crate signal_hook;
 
 #[cfg(test)]
 #[allow(dead_code)]
@@ -37,7 +36,7 @@ mod session_types_tests {
         connect(server, drop_client);
     }
 
-    #[test]
+    // #[test]
     fn server_incomplete_segfaults() {
         connect(drop_server, client);
     }

### 2019-03-21 tests/session_types_tests.rs
Fixed import statements

@@ -11,7 +11,6 @@ mod session_types_tests {
     use std::error::Error;
     use std::thread;
     use session_types::*;
-    use signal_hook::{iterator::Signals, SIGSEGV};
 
     fn client(c: Chan<(), Send<(), Eps>>) {
         c.send(()).close();

### 2019-03-23 tests/session_types_tests.rs
Moved tests to their own file; removed tests for session-types.

@@ -1,47 +1,0 @@
-#![feature(test)]
-
-extern crate session_types;
-extern crate test;
-
-#[cfg(test)]
-#[allow(dead_code)]
-mod session_types_tests {
-
-    use std::boxed::Box;
-    use std::error::Error;
-    use std::thread;
-    use session_types::*;
-
-    fn client(c: Chan<(), Send<(), Eps>>) {
-        c.send(()).close();
-    }
-
-    fn server(c: Chan<(), Recv<(), Eps>>) {
-        let (c, ()) = c.recv();
-        c.close();
-    }
-
-    fn drop_client(_c: Chan<(), Send<(), Eps>>) {}
-
-    fn drop_server(_c: Chan<(), Recv<(), Eps>>) {}
-
-    #[test]
-    fn server_client_works() {
-        connect(server, client);
-    }
-
-    // #[test]
-    fn client_incomplete_panics() {
-        connect(server, drop_client);
-    }
-
-    // #[test]
-    fn server_incomplete_segfaults() {
-        connect(drop_server, client);
-    }
-
-    // #[test]
-    fn server_client_incomplete_panics() {
-        connect(drop_server, drop_client);
-    }
-}

### 2019-03-23 tests/tests.rs

@@ -1,0 +1,348 @@
+extern crate rand;
+extern crate rusty_variation;
+
+use rand::{Rng, thread_rng};
+use rusty_variation::*;
+use std::boxed::Box;
+use std::error::Error;
+use std::marker;
+use std::mem;
+use std::sync::mpsc;
+
+
+// Test sending a ping across threads.
+
+#[test]
+fn ping_works() {
+    assert!(|| -> Result<(), Box<Error>> {
+
+        let s = fork!(move |s: Send<(), End>| {
+            let s = send((), s)?;
+            close(s)
+        });
+        let ((), s) = recv(s)?;
+        close(s)
+
+    }().is_ok());
+}
+
+
+// Test a simple calculator server, implemented using binary choice.
+
+type NegServer<N> = Recv<N, Send<N, End>>;
+type NegClient<N> = <NegServer<N> as Session>::Dual;
+
+type AddServer<N> = Recv<N, Recv<N, Send<N, End>>>;
+type AddClient<N> = <AddServer<N> as Session>::Dual;
+
+type SimpleCalcServer<N> = Offer<NegServer<N>, AddServer<N>>;
+type SimpleCalcClient<N> = <SimpleCalcServer<N> as Session>::Dual;
+
+fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<Error>> {
+    offer_either(s,
+                 |s: NegServer<i32>| {
+                     let (x, s) = recv(s)?;
+                     let s = send(-x, s)?;
+                     close(s)
+                 },
+                 |s: AddServer<i32>| {
+                     let (x, s) = recv(s)?;
+                     let (y, s) = recv(s)?;
+                     let s = send(x.wrapping_add(y), s)?;
+                     close(s)
+                 })
+}
+
+#[test]
+fn simple_calc_works() {
+    assert!(|| -> Result<(), Box<Error>> {
+
+        let mut rng = thread_rng();
+
+        // Test the negation function.
+        {
+            let s: SimpleCalcClient<i32> = fork!(simple_calc_server);
+            let x: i32 = rng.gen();
+            let s = choose_left::<_, AddClient<i32>>(s)?;
+            let s = send(x, s)?;
+            let (y, End) = recv(s)?;
+            assert_eq!(-x, y);
+        }
+
+        // Test the addition function.
+        {
[…]

### 2019-04-07 tests/tests.rs
Added preamble, wrote out type system for RV, fixed bug.

@@ -330,19 +330,19 @@ fn deadlock_forget() {
 }
 
 
-// #[allow(dead_code)]
-// fn deadlock_new() {
-//
-//     let (s1, r1) = <Send<(), End>>::new();
-//     let r2 = fork(move |s2: Send<(), End>| {
-//         let (x, End) = recv(r1)?;
-//         let End = send(x, s2)?;
-//         Ok(())
-//     });
-//
-//     || -> Result<(), Box<Error>> {
-//         let (x, End) = recv(r2)?;
-//         let End = send(x, s1)?;
-//         Ok(())
-//     }().unwrap();
-// }
+#[allow(dead_code)]
+fn deadlock_new() {
+
+    let (s1, r1) = <Send<(), End>>::new();
+    let r2 = fork(move |s2: Send<(), End>| {
+        let (x, End) = recv(r1)?;
+        let End = send(x, s2)?;
+        Ok(())
+    });
+
+    || -> Result<(), Box<Error>> {
+        let (x, End) = recv(r2)?;
+        let End = send(x, s1)?;
+        Ok(())
+    }().unwrap();
+}

### 2019-04-07 src/lib.rs
Added preamble, wrote out type system for RV, fixed bug.

@@ -5,8 +5,8 @@ use std::boxed::Box;
 use std::error::Error;
 use std::marker;
 use std::mem;
-use std::thread::{JoinHandle,spawn};
-use crossbeam_channel::{Sender, Receiver, bounded};
+use std::thread::{JoinHandle, spawn};
+use crossbeam_channel::{Sender, Receiver, Select, bounded};
 use either::Either;
 
 /// The session types supported.
@@ -60,7 +60,7 @@ impl<T: marker::Send, S: Session> Session for Recv<T, S> {
 }
 
 
-/// The communication primitives.
+// Send and receive
 
 pub fn send<'a, T, S>(x: T, s: Send<T, S>) -> Result<S, Box<Error + 'a>>
 where
@@ -81,20 +81,27 @@ where
     Ok((v, s))
 }
 
-pub fn close(s: End) -> Result<(), Box<Error>> {
-    let End = s;
-    Ok(())
-}
+
+
+// Cancellation and closing
 
 pub fn cancel<T>(x: T) -> Result<(), Box<Error>> {
     mem::drop(x);
     Ok(())
 }
 
+pub fn close(s: End) -> Result<(), Box<Error>> {
+    let End = s;
+    Ok(())
+}
+
+
+// Fork
+
 pub fn fork_with_thread_id<S, P>(p: P) -> (JoinHandle<()>, S::Dual)
 where
     S: Session + 'static,
-    P: Fn(S) -> Result<(), Box<Error>> + marker::Send + 'static
+    P: FnOnce(S) -> Result<(), Box<Error>> + marker::Send + 'static
 {
     let (there, here) = Session::new();
     let other_thread = spawn(move || {
@@ -109,15 +116,19 @@ where
 pub fn fork<S, P>(p: P) -> S::Dual
 where
     S: Session + 'static,
-    P: Fn(S) -> Result<(), Box<Error>> + marker::Send + 'static
+    P: FnOnce(S) -> Result<(), Box<Error>> + marker::Send + 'static
 {
     fork_with_thread_id(p).1
 }
+
+
+// Binary choice
 
 pub type Offer<S1, S2> =
     Recv<Either<S1, S2>, End>;
 pub type Choose<S1, S2> =
     Send<Either<<S1 as Session>::Dual, <S2 as Session>::Dual>, End>;
+
 
 pub fn offer_either<'a, S1, S2, F, G, R>(s: Offer<S1, S2>, f: F, g: G)
[…]

### 2019-04-10 src/lib.rs
Switched to synchronous close; for old behaviour cancel End

@@ -12,7 +12,10 @@ use either::Either;
 /// The session types supported.
 
 #[derive(Debug)]
-pub struct End;
+pub struct End {
+    sender: Sender<()>,
+    receiver: Receiver<()>,
+}
 
 #[must_use]
 #[derive(Debug)]
@@ -37,7 +40,11 @@ impl Session for End {
     type Dual = End;
 
     fn new() -> (Self, Self::Dual) {
-        return (End, End);
+        let (sender1, receiver1) = bounded::<()>(1);
+        let (sender2, receiver2) = bounded::<()>(1);
+
+        return (End { sender: sender1, receiver: receiver2 },
+                End { sender: sender2, receiver: receiver1 });
     }
 }
 
@@ -91,7 +98,8 @@ pub fn cancel<T>(x: T) -> Result<(), Box<Error>> {
 }
 
 pub fn close(s: End) -> Result<(), Box<Error>> {
-    let End = s;
+    s.sender.send(())?;
+    s.receiver.recv()?;
     Ok(())
 }
 
@@ -138,7 +146,8 @@ where
     F: FnOnce(S1) -> Result<R, Box<Error + 'a>>,
     G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
 {
-    let (e, End) = recv(s)?;
+    let (e, s) = recv(s)?;
+    close(s)?;
     e.either(f, g)
 }
 
@@ -148,7 +157,8 @@ where
     S2: Session + 'a,
 {
     let (here, there) = S1::new();
-    let End = send(Either::Left(there), s)?;
+    let s = send(Either::Left(there), s)?;
+    close(s)?;
     Ok(here)
 }
 
@@ -158,7 +168,8 @@ where
     S2: Session + 'a,
 {
     let (here, there) = S2::new();
-    let End = send(Either::Right(there), s)?;
+    let s = send(Either::Right(there), s)?;
+    close(s)?;
     Ok(here)
 }
 

### 2019-04-10 tests/tests.rs

@@ -65,7 +65,8 @@ fn simple_calc_works() {
             let x: i32 = rng.gen();
             let s = choose_left::<_, AddClient<i32>>(s)?;
             let s = send(x, s)?;
-            let (y, End) = recv(s)?;
+            let (y, s) = recv(s)?;
+            close(s)?;
             assert_eq!(-x, y);
         }
 
@@ -77,7 +78,8 @@ fn simple_calc_works() {
             let s = choose_right::<NegClient<i32>, _>(s)?;
             let s = send(x, s)?;
             let s = send(y, s)?;
-            let (z, End) = recv(s)?;
+            let (z, s) = recv(s)?;
+            close(s)?;
             assert_eq!(x.wrapping_add(y), z);
         }
 
@@ -158,8 +160,7 @@ fn cancel_recv_works() {
 
     assert!(|| -> Result<(), Box<Error>> {
 
-        cancel(s)?;
-        Ok(())
+        cancel(s)
 
     }().is_ok());
 
@@ -170,14 +171,16 @@ fn cancel_recv_works() {
 fn cancel_send_works() {
 
     let (other_thread, s) = fork_with_thread_id(
-        move |s: Recv<(), End>| {cancel(s)});
+        move |s: Recv<(), End>| {
+            cancel(s)
+        });
 
     assert!(|| -> Result<(), Box<Error>> {
 
-        send((), s)?;
-        Ok(())
+        let s = send((), s)?;
+        close(s)
 
-    }().is_ok());
+    }().is_err());
 
     assert!(other_thread.join().is_ok());
 }
@@ -194,10 +197,9 @@ fn delegation_works() {
     assert!(|| -> Result<(), Box<Error>> {
 
         let u = send(s, u)?;
-        close(u)?;
-        Ok(())
+        close(u)
 
-    }().is_ok());
+    }().is_err());
 
     assert!(other_thread1.join().is_err());
     assert!(other_thread2.join().is_ok());
@@ -304,13 +306,13 @@ fn deadlock_loop() {
             // Let's trick the reachability checker
             if false { break; }
         }
-        let End = send((), s)?;
-        Ok(())
+        let s = send((), s)?;
+        close(s)
     });
 
     || -> Result<(), Box<Error>> {
-        let ((), End) = recv(s)?;
-        Ok(())
+        let ((), s) = recv(s)?;
+        close(s)
[…]

### 2019-04-11 src/lib.rs
Removed superfluous error annotations

@@ -6,7 +6,7 @@ use std::error::Error;
 use std::marker;
 use std::mem;
 use std::thread::{JoinHandle, spawn};
-use crossbeam_channel::{Sender, Receiver, Select, bounded};
+use crossbeam_channel::{Sender, Receiver, bounded};
 use either::Either;
 
 /// The session types supported.
@@ -69,14 +69,14 @@ impl<T: marker::Send, S: Session> Session for Recv<T, S> {
 
 // Send and receive
 
-pub fn send<'a, T, S>(x: T, s: Send<T, S>) -> Result<S, Box<Error + 'a>>
+pub fn send<'a, T, S>(x: T, s: Send<T, S>) -> S
 where
     T: marker::Send + 'a,
     S: Session + 'a,
 {
     let (here, there) = S::new();
     s.channel.send((x, there)).unwrap_or(());
-    Ok(here)
+    here
 }
 
 pub fn recv<'a, T, S>(s: Recv<T, S>) -> Result<(T, S), Box<Error + 'a>>
@@ -92,15 +92,13 @@ where
 
 // Cancellation and closing
 
-pub fn cancel<T>(x: T) -> Result<(), Box<Error>> {
+pub fn cancel<T>(x: T) -> () {
     mem::drop(x);
-    Ok(())
 }
 
-pub fn close(s: End) -> Result<(), Box<Error>> {
-    s.sender.send(())?;
-    s.receiver.recv()?;
-    Ok(())
+pub fn close(s: End) -> () {
+    s.sender.send(()).unwrap_or(());
+    s.receiver.recv().unwrap_or(());
 }
 
 
@@ -147,30 +145,30 @@ where
     G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
 {
     let (e, s) = recv(s)?;
-    close(s)?;
+    close(s);
     e.either(f, g)
 }
 
-pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> Result<S1, Box<Error + 'a>>
+pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> S1
 where
     S1: Session + 'a,
     S2: Session + 'a,
 {
     let (here, there) = S1::new();
-    let s = send(Either::Left(there), s)?;
-    close(s)?;
-    Ok(here)
+    let s = send(Either::Left(there), s);
+    close(s);
+    here
 }
 
[…]

### 2019-04-11 tests/tests.rs

@@ -17,11 +17,13 @@ fn ping_works() {
     assert!(|| -> Result<(), Box<Error>> {
 
         let s = fork(move |s: Send<(), End>| {
-            let s = send((), s)?;
-            close(s)
+            let s = send((), s);
+            close(s);
+            Ok(())
         });
         let ((), s) = recv(s)?;
-        close(s)
+        close(s);
+        Ok(())
 
     }().is_ok());
 }
@@ -42,14 +44,16 @@ fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<Error>> {
     offer_either(s,
                  |s: NegServer<i32>| {
                      let (x, s) = recv(s)?;
-                     let s = send(-x, s)?;
-                     close(s)
+                     let s = send(-x, s);
+                     close(s);
+                     Ok(())
                  },
                  |s: AddServer<i32>| {
                      let (x, s) = recv(s)?;
                      let (y, s) = recv(s)?;
-                     let s = send(x.wrapping_add(y), s)?;
-                     close(s)
+                     let s = send(x.wrapping_add(y), s);
+                     close(s);
+                     Ok(())
                  })
 }
 
@@ -63,10 +67,10 @@ fn simple_calc_works() {
         {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
-            let s = choose_left::<_, AddClient<i32>>(s)?;
-            let s = send(x, s)?;
+            let s = choose_left::<_, AddClient<i32>>(s);
+            let s = send(x, s);
             let (y, s) = recv(s)?;
-            close(s)?;
+            close(s);
             assert_eq!(-x, y);
         }
 
@@ -75,11 +79,11 @@ fn simple_calc_works() {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
             let y: i32 = rng.gen();
-            let s = choose_right::<NegClient<i32>, _>(s)?;
-            let s = send(x, s)?;
-            let s = send(y, s)?;
+            let s = choose_right::<NegClient<i32>, _>(s);
+            let s = send(x, s);
+            let s = send(y, s);
             let (z, s) = recv(s)?;
-            close(s)?;
+            close(s);
[…]

### 2019-04-12 tests/benchmark_tests.rs
Fixed benchmark tests

@@ -30,28 +30,29 @@ mod rusty_variation_bench {
     fn calc_server(s: CalcSrv) -> Result<(), Box<Error>> {
         ::rusty_variation::offer!(s, {
             CalcOp::CLOSE(s) => {
-                close(s)
+                close(s);
+                Ok(())
             },
             CalcOp::ADD(s) => {
                 let (x, s) = recv(s)?;
                 let (y, s) = recv(s)?;
-                let s = send(x + y, s)?;
+                let s = send(x + y, s);
                 calc_server(s)
             },
             CalcOp::NEGATE(s) => {
                 let (x, s) = recv(s)?;
-                let s = send(-x, s)?;
+                let s = send(-x, s);
                 calc_server(s)
             },
             CalcOp::SQRT(s) => {
                 let (x, s) = recv(s)?;
-                let s = send(x.sqrt(), s)?;
+                let s = send(x.sqrt(), s);
                 calc_server(s)
             },
             CalcOp::EVAL(s) => {
                 let (f, s) = recv(s)?;
                 let (x, s) = recv(s)?;
-                let s = send(f(x), s)?;
+                let s = send(f(x), s);
                 calc_server(s)
             },
         })
@@ -59,12 +60,13 @@ mod rusty_variation_bench {
 
     fn neg_client(s: CalcCli) -> Result<(), Box<Error>> {
         let n = thread_rng().gen();
-        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s)?;
-        let s = send(n, s)?;
+        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s);
+        let s = send(n, s);
         let (n_, s) = recv(s)?;
         assert_eq!(-n, n_);
-        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s)?;
-        close(s)
+        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s);
+        close(s);
+        Ok(())
     }
 
     #[bench]

### 2019-04-12 tests/tests.rs
Fixed tests

@@ -188,7 +188,7 @@ fn cancel_send_works() {
         close(s);
         Ok(())
 
-    }().is_err());
+    }().is_ok());
 
     assert!(other_thread.join().is_ok());
 }
@@ -211,7 +211,7 @@ fn delegation_works() {
         close(u);
         Ok(())
 
-    }().is_err());
+    }().is_ok());
 
     assert!(other_thread1.join().is_err());
     assert!(other_thread2.join().is_ok());

### 2019-04-12 src/lib.rs
Suppress panic message in child threads

@@ -6,6 +6,7 @@ use std::error::Error;
 use std::marker;
 use std::mem;
 use std::thread::{JoinHandle, spawn};
+use std::panic;
 use crossbeam_channel::{Sender, Receiver, bounded};
 use either::Either;
 
@@ -111,6 +112,9 @@ where
 {
     let (there, here) = Session::new();
     let other_thread = spawn(move || {
+        panic::set_hook(Box::new(|_info| {
+            // do nothing
+        }));
         match p(there) {
             Ok(()) => (),
             Err(e) => panic!("{}", e.description()),

### 2019-04-17 src/lib.rs
Fixed tests; allow for failing close

@@ -12,6 +12,7 @@ use either::Either;
 
 /// The session types supported.
 
+#[must_use]
 #[derive(Debug)]
 pub struct End {
     sender: Sender<()>,
@@ -97,9 +98,10 @@ pub fn cancel<T>(x: T) -> () {
     mem::drop(x);
 }
 
-pub fn close(s: End) -> () {
+pub fn close(s: End) -> Result<(), Box<Error>> {
     s.sender.send(()).unwrap_or(());
-    s.receiver.recv().unwrap_or(());
+    s.receiver.recv()?;
+    Ok(())
 }
 
 
@@ -149,30 +151,30 @@ where
     G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
 {
     let (e, s) = recv(s)?;
-    close(s);
+    close(s)?;
     e.either(f, g)
 }
 
-pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> S1
+pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> Result<S1, Box<Error>>
 where
     S1: Session + 'a,
     S2: Session + 'a,
 {
     let (here, there) = S1::new();
     let s = send(Either::Left(there), s);
-    close(s);
-    here
+    close(s)?;
+    Ok(here)
 }
 
-pub fn choose_right<'a, S1, S2>(s: Choose<S1, S2>) -> S2
+pub fn choose_right<'a, S1, S2>(s: Choose<S1, S2>) -> Result<S2, Box<Error>>
 where
     S1: Session + 'a,
     S2: Session + 'a,
 {
     let (here, there) = S2::new();
     let s = send(Either::Right(there), s);
-    close(s);
-    here
+    close(s)?;
+    Ok(here)
 }
 
 
@@ -183,7 +185,7 @@ macro_rules! offer {
     ($session:expr, { $($pat:pat => $result:expr,)* }) => {
         (move || -> Result<_, _> {
             let (l, s) = recv($session)?;
-            close(s);
+            close(s)?;
             match l {
                 $(
                     $pat => $result,
@@ -195,12 +197,16 @@ macro_rules! offer {
 
 #[macro_export]
 macro_rules! choose {
-    ($label:path, $session:expr) => {{
-        let (here, there) = <_ as Session>::new();
-        let s = send($label(there), $session);
-        close(s);
-        here
-    }};
+    ($label:path, $session:expr) => {
+        (move || -> Result<_, _> {
[…]

### 2019-04-17 tests/benchmark_tests.rs

@@ -30,7 +30,7 @@ mod rusty_variation_bench {
     fn calc_server(s: CalcSrv) -> Result<(), Box<Error>> {
         ::rusty_variation::offer!(s, {
             CalcOp::CLOSE(s) => {
-                close(s);
+                close(s)?;
                 Ok(())
             },
             CalcOp::ADD(s) => {
@@ -60,12 +60,12 @@ mod rusty_variation_bench {
 
     fn neg_client(s: CalcCli) -> Result<(), Box<Error>> {
         let n = thread_rng().gen();
-        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s);
+        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s)?;
         let s = send(n, s);
         let (n_, s) = recv(s)?;
         assert_eq!(-n, n_);
-        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s);
-        close(s);
+        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s)?;
+        close(s)?;
         Ok(())
     }
 

### 2019-04-17 tests/tests.rs

@@ -18,11 +18,11 @@ fn ping_works() {
 
         let s = fork(move |s: Send<(), End>| {
             let s = send((), s);
-            close(s);
+            close(s)?;
             Ok(())
         });
         let ((), s) = recv(s)?;
-        close(s);
+        close(s)?;
         Ok(())
 
     }().is_ok());
@@ -45,14 +45,14 @@ fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<Error>> {
                  |s: NegServer<i32>| {
                      let (x, s) = recv(s)?;
                      let s = send(-x, s);
-                     close(s);
+                     close(s)?;
                      Ok(())
                  },
                  |s: AddServer<i32>| {
                      let (x, s) = recv(s)?;
                      let (y, s) = recv(s)?;
                      let s = send(x.wrapping_add(y), s);
-                     close(s);
+                     close(s)?;
                      Ok(())
                  })
 }
@@ -67,10 +67,10 @@ fn simple_calc_works() {
         {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
-            let s = choose_left::<_, AddClient<i32>>(s);
+            let s = choose_left::<_, AddClient<i32>>(s)?;
             let s = send(x, s);
             let (y, s) = recv(s)?;
-            close(s);
+            close(s)?;
             assert_eq!(-x, y);
         }
 
@@ -79,11 +79,11 @@ fn simple_calc_works() {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
             let y: i32 = rng.gen();
-            let s = choose_right::<NegClient<i32>, _>(s);
+            let s = choose_right::<NegClient<i32>, _>(s)?;
             let s = send(x, s);
             let s = send(y, s);
             let (z, s) = recv(s)?;
-            close(s);
+            close(s)?;
             assert_eq!(x.wrapping_add(y), z);
         }
 
@@ -107,14 +107,14 @@ fn nice_calc_server(s: NiceCalcServer<i32>) -> Result<(), Box<Error>> {
         CalcOp::Neg(s) => {
             let (x, s) = recv(s)?;
             let s = send(-x, s);
-            close(s);
+            close(s)?;
             Ok(())
         },
[…]

### 2019-05-05 doc/abstract/main.tex
Worked on paper

\documentclass[sigconf,natbib=false]{acmart}
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\pagestyle{plain} % removes running headers
\input{preamble}

\begin{document}
\title{Rusty Variation}
\subtitle{Deadlock-free Sessions with Failure in Rust}
\author{%
% Wen Kokke
}
\affiliation{%
%  LFCS, University of Edinburgh, 10 Crichton St,
%  EH8 9AB, Edinburgh, United Kingdom
}
\email{%
% wen.kokke@ed.ac.uk
}

\begin{abstract}
  Rusty Variation (RV) is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees. Programs written using RV are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
\end{abstract}

\maketitle

\section{Problem \& Motivation}\label{sec:motivation}
In concurrent programming, processes commonly exchange information by sending messages on shared channels. However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.

Session types are a substructural typing discipline, introduced by \textcite{honda1993}, which describe communication protocols as \emph{channel types}, capturing both the type and the order of messages. Session type systems, then, ensure that programs correctly implement specified protocols.

Rust is an immensely popular systems programming language.\footnote{\url{https://insights.stackoverflow.com/survey/2019/\#technology-\_-most-loved-dreaded-and-wanted-languages}} It aims to combine efficiency with abstraction and thread and memory safety. However, out of the box, it does not offer a way to specify protocols and verify communication.

In this paper, we present Rusty Variation\footnote{\url{https://github.com/reviewer2/rusty-variation/}}, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and races.
We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}.\footnote{\url{https://doc.rust-lang.org/nightly/book/}}

\section{Background \& Related Work}\label{sec:background}
\Textcite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by \textcite{gay2009}. Their implementation guarantees that programs correctly implement specified communication protocols with some caveats. LAST depends upon linear types, meaning that sessions have to be used exactly once. Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped. The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}. In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \lstinline{#[must_use]} annotation. However, in general, this is not something the compiler can guarantee. Furthermore, LAST ignores the possibility of communication failures, such as network errors, which cannot be detected at compile-time, yet can cause a channel to be dropped.

We embrace the affine, and base our work on Exceptional GV by \textcite[EGV]{fowler2019}. Exceptional GV is a descendant of Good Variation~\parencite[GV]{wadler2012}, which itself is a descendant of LAST. Crucially, EGV has an affine type system, and allows for the explicit cancellation of sessions. Furthermore, EGV offers a stronger metatheory than LAST. In addition to session fidelity, EGV also guarantees global progress, freedom from deadlocks and livelocks, confluence, and termination. That is to say, there are no locks or race conditions, and all programs eventually halt. For a discussion of EGV and its metatheory, we refer the reader to the work by \textcite{fowler2019}.

\section{Contribution: Rusty Variation}\label{sec:contribution}
Our contribution, Rusty Variation, is an implementation of EGV in Rust. We claim that it is a faithful implementation, which preserves much of EGV's metatheory---we go further into this in \cref{sec:verification}. However, we depart from it in two ways:

First, EGV relies upon exceptions and handlers. Rust, however, has rejected exceptions in favour of monadic errors, \ie, wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.

Second, the core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee session fidelity, livelock freedom, or---well---termination. The work by \textcite{jespersen2015} also suffers from this shortcoming. We can demonstrate this by writing the program which loops forever, and then sends a message. The Rust type checker does not see a problem with this, but obviously the message will never be sent. However, we claim that if the program terminates, it respected the communication protocol.

\subsection{Sessions and duality}
Any type built using the primitive types below is a valid session:
\begin{lstlisting}
pub struct End { ... }
pub struct Send<T, S: Session> { ... }
pub struct Recv<T, S: Session> { ... }
\end{lstlisting}
Session \emph{types} are public, but their members are not. Therefore, users cannot construct their own misbehaving sessions. The continuation of a session must itself be a session. This is enforced by the \lstinline{Session} constraint. In addition to this ``kinding'' of session types, the \lstinline{Session} trait also implements duality and session generation:
\begin{lstlisting}
pub trait Session {
  type Dual: Session<Dual=Self>;
  fn new() -> (Self, Self::Dual);
}
\end{lstlisting}
Duality works as expected: \lstinline{Send<T, S>} is dual to \lstinline{Recv<T, S::Dual>}---and vice versa---and \lstinline{End} is self-dual. The constraint \lstinline{Session<Dual=Self>} enforces that the dual of a session type is a session type , and that \lstinline{S::Dual::Dual} is equal to \lstinline{S} (involutivity).

The \lstinline{new} function generates a new session-typed channel and returns two \emph{dual} channel endpoints. Unlike the rest of the functions discussed in this section, \lstinline{new} is \emph{not} part of the public interface---see \cref{sec:limitations}.

Using these types, we can define the types of servers which offer negation and addition as follows, where \lstinline{i32} is the type of 32-bit integers:
\begin{lstlisting}
type NegSrv = Recv<i32, Send<i32, End>>;
type AddSrv = Recv<i32, Recv<i32, Send<i32, End>>>;
\end{lstlisting}
The client types are obtained by duality.

\subsection{Send, receive, close, and fork}\label{sec:basics}
The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. Sending is non-blocking, and never fails. Sending a value on a cancelled channel is equivalent to cancelling that value---see \cref{sec:cancellation}. Receiving is blocking, and fails if the channel is cancelled. They have the following types:
\begin{lstlisting}
pub fn send<T, S: Session>(x: T, s: Send<T, S>) -> S
pub fn recv<T, S: Session>(s: Recv<T, S) -> Option<(T, S)>
\end{lstlisting}
The \lstinline{close} function closes a \emph{completed} session. Closing is \emph{synchronous}, meaning that \lstinline{close} blocks until the dual end point is also closed. Closing a cancelled channel fails. It has the following type:
\begin{lstlisting}
pub fn close(s: End) -> Option<()>
\end{lstlisting}
The \lstinline{fork} function creates a new session, and spawns a new thread. The new thread runs the first argument of \lstinline{fork}---a code block---which receives one endpoint, and \lstinline{fork} returns the dual endpoint. It has the following type (where \lstinline{FnOnce(S) -> Option<()>} is the type of functions from \lstinline{S} to \lstinline{Option<()>} which can be invoked only once):
\begin{lstlisting}
pub fn fork<S: Session>(p: P) -> S::Dual;
where
  P: FnOnce(S) -> Option<()>,
\end{lstlisting}
The code block passed to \lstinline{fork} is allowed to fail. If it does, the child thread panics silently, and all open sessions are cancelled.

Now that we have covered the basics, let us consider an example. The following program forks off a process which sends a ``ping'' along the channel, then waits to receive that ping and returns:
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  let s = send((), s);
  close(s)
});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
The \lstinline{let x = M?; N} construct is Rust's ``monadic bind'' notation for programs which may return errors. If \lstinline{recv(s)} succeeds, the \lstinline{?}-operator unpacks the \lstinline{Option}. If \lstinline{recv(s)} fails, the \lstinline{?}-operator short-circuits, skips the rest of the statements, and returns the error.

\subsection{Cancellation}\label{sec:cancellation}
The \lstinline{cancel} function drops a value and cancels all sessions \emph{and} their duals referenced in that value. This cancellation propagates: any process attempting to receive on or close a cancelled channel fails and cancels all their open sessions; any value sent on a cancelled channel is also cancelled. This ensures that no process is left waiting for a message that will never come.
\begin{lstlisting}
pub fn cancel<T>(x: T) -> ()
\end{lstlisting}
Using \lstinline{cancel}, we can implement the following program. This program forks off a process which is \emph{expected} to send a ``ping''. However, instead the forked process cancels the session. Therefore, the call to \lstinline{recv} fails, the \lstinline{?}-operator short-circuits, and the whole program returns an error:
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  cancel(s);
  Some(())
});
let (z, s) = recv(s)?;
close(s)
\end{lstlisting}
The cancellation mechanism is implemented through destructors. In the current implementation, cancellation is inherited from the implementation of channels in the \lstinline{crossbeam_channel} crate.

The \lstinline{End}, \lstinline{Send}, and \lstinline{Recv} structures are annotated with \lstinline{#[must_use]}, to encourage the user to complete the protocol or cancel it explicitly. However, should a session be dropped, this is interpreted as \emph{implicit} cancellation. This ensures deadlock freedom, even in Rust's affine type system.

\subsection{Branching with offer! and choose!}
The \lstinline{choose_left}, \lstinline{choose_right}, and \lstinline{offer_either} functions let processes offer or make a \emph{binary} choice. These are \emph{derived} constructs, in that they are implemented using the primitive constructs~\parencite{honda1993,dardha2017}: they send or receive a value of the sum type, \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.
\begin{lstlisting}
type Choose<S1, S2> = Send<Either<S1, S2>, End>;
type Offer<S1, S2> = Recv<Either<S1, S2>, End>;

pub fn choose_left<S1, S2>(s: Choose<S1, S2>) -> S1
where
  S1: Session, S2: Session,

pub fn choose_right<S1, S2>(s: Choose<S1, S2>) -> S2
where
  S1: Session, S2: Session,

pub fn offer_either<S1, S2, P1, P2, R>(
       s: Offer<S1, S2>, p1: P1, p2: P2) -> Option<R>
where
  S1: Session, S2: Session,
  P1: FnOnce(S1) -> Option<R>,
  P2: FnOnce(S2) -> Option<R>,
\end{lstlisting}
While these functions suffice to encode choice, they are not useful in practice---a protocol with a four-way choice would require three nested \lstinline{Offer} constructs, and several calls to \lstinline{offer_either}. Code written in this style rapidly becomes unreadable. For this reason, we also define two macros, \lstinline{choose!} and \lstinline{offer!}, which generalise binary choice using \lstinline{Either} to a choice on \emph{any} \lstinline{enum} where each value wraps a session. This allows processes to offer and make choices between many \emph{labelled} branches, and allows us to define the type for a calculator server as:
\begin{lstlisting}
type CalcSrv = Recv<CalcOp, End>;
enum CalcOp { Neg(NegSrv), Add(AddSrv) }
\end{lstlisting}
Using the \lstinline{choose!} and \lstinline{offer!} macros, we can write the following program. This program forks off a calculator server, which offers the choice between negation and addition, selects addition, sends some numbers, and then returns the result:
\begin{lstlisting}
let s = fork(move |s: CalcSrv| {
  offer!(s, {
    CalcOp::Neg(s) => {
      let (x, s) = recv(s)?;
      let s = send(-x, s);
      close(s)
    },
    CalcOp::Add(s) => {
      let (x, s) = recv(s)?;
      let (y, s) = recv(s)?;
      let s = send(x + y, s);
      close(s)
    },
  })});
let s = choose!(CalcOp::Add, s);
let s = send(4, s);
let s = send(5, s);
let (z, s) = recv(s)?;
close(s)?;
Some(z)
\end{lstlisting}

\subsection{Implementation}
We use the asynchronous channels from \lstinline{crossbeam_channel}. These are ideal, as they already implement the correct cancellation behaviour. Channels are typed, and can only transfer values of a single type. Therefore, we use one-shot channels, and encode sessions following \textcite{tart2017,scalas2016,padovani2017}: each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value, and each \lstinline{recv} receives a value together with a channel on which to continue the session. Fortunately, the channels in \lstinline{crossbeam_channel} are highly optimised for one-shot usage.

\subsection{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Rusty Variation deadlock. The first is by looping, which we discussed earlier. The other two are explicit, and surprisingly, \lstinline{unsafe} is not one of them.

The biggest issue lies in the fact that Rust's semantics do not guarantee that destructors are actually run. This is important for, \eg, doubly linked lists, and other data structures which require manually managed memory. For this reason, it is always possible to ``forget'' a value, using \lstinline{std::mem::forget}, without running its destructors. The following simple program deadlocks:
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  std::mem::forget(s);
  Some(())
});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
Second, it is not possible---at the time of writing---to export a trait without exporting all its members. Therefore, any user can use the \lstinline{new} function, and doing so, it is trivial to write a program which deadlocks: 
\begin{lstlisting}
let (s1, r1) = Send<Void, End>::new();
let (s2, r2) = Send<Void, End>::new();
std::thread::spawn(move || {
  let (v, r1) = recv(r1)?;
  close(r1)?;
  let s2 = send(v, s2);
  close(s2)?;
  Some(())
});
let (v, r2) = recv(r2)?;
close(r2)?;
let s1 = send(v, s1);
close(s1)?;
\end{lstlisting}

\subsection{Future Work: Testing \& Verification}\label{sec:verification}
In \cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.

Firstly, RV is unit tested. We test various use-cases from EGV, from basic operations and their interactions with cancellation, to delegation and recursive session types. This test suite has caught several errors in the implementation. However, unit tests are error prone, and it is hard (if not impossible) to obtain a set of tests which cover all use-cases of a library. Unfortunately, tools such as RustBelt~\parencite{jung2017} and Rust Distilled~\parencite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of RV.

In the absence of a mature formal semantics for Rust, we are designing a formal language, RV (or ``formal RV''), which matches, as close as possible, the semantics of our Rust library. It is a concurrent lambda calculus with heap-based shared-memory semantics. We plan to prove that formal RV guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom. Furthermore, we plan to show that there exists a translation from EGV to formal RV, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\parencite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV. This would show that formal RV is a faithful implementation of EGV.

Lastly, we plan to ensure the correspondence between formal RV and the implementation of RV, using property-based testing. We plan to generate a series of random formal RV terms, using Neat~\parencite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal RV.

\printbibliography
\end{document}

### 2019-05-05 doc/paper2019/proofs.tex
Worked on paper

\begin{lemma}\label{lem:translate-term}
  If $\tm{M}\gvRedArrPure\tm{M'}$,
  then $\tm{\ftm{M}}\gvRedArrPure^{+}\tm{\ftm{M'}}$.
\end{lemma}
\begin{proof}
  By induction on the reduction, inserting additional reductions for error handling.
\end{proof}

\begin{lemma}
  If $\tm{\ftm{M}}\gvRedArrPure\tm{N}$,
  then there exists a term $\tm{M'}$ such that
  $\tm{M}\gvRedArrPure\tm{M'}$ and
  $\tm{N}\gvRedArrPure^{\star}\tm{\ftm{M'}}$.
\end{lemma}
\begin{proof}
  By induction on the reduction, inserting additional reductions for error handling.
\end{proof}

\begin{lemma}\label{lem:translate-raise}
  $\ftm{\gvPlug{P}{\gvRaise}}\gvRedArrPure^{\star}\rvNone$
\end{lemma}
\begin{proof}
  By induction on $\tm{P}$. Apply \textsc{E-None}.
\end{proof}

\begin{theorem}
  If $\tm{\gvConf{C}}$ is well-typed and $\tm{\gvConf{C}}\gvRedArr\tm{\gvConf{C'}}$,
  then \\ $\tm{\ftm{\gvConf{C}}}\gvRedArr^{+}\tm{\ftm{\gvConf{C'}}}$.
\end{theorem}
\begin{proof}
  By case analysis on the reduction.
  We handle applications of \textsc{E-LiftC} in each case as a prefix $\tm{\gvConf{G}}$.
  \begin{itemize}
  \item
    Case \textsc{E-Cancel}.
    We have $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvPlug{\gvConf{F}}{\gvCancel{a}}}$. As $\tm{\gvConf{C}}$ is well-typed, $\gvVar{a}$ does not occur in $\tm{\gvConf{G}}$ or $\tm{\gvConf{F}}$, and $\ren{\gvConf{C}}(\gvVar{a})$ occurs at most once in $\ftm{\gvConf{C}}$, \ie $\ftmArc{a}{\gvConf{C}}\le{1}$. \\
    Apply \textsc{E-Zap}.
  \item
    Case \textsc{E-Zap}.
    We have $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvPar{\gvZap{a}}{\gvBuf{a}{\gvVar{U}\gvVecAdd\gvVec{V}}{b}{\gvVec{W}}}}$. For each channel $\gvVar{c}$ in $\gvVar{U}$: As $\tm{\gvConf{C}}$ is well-typed, $\gvVar{c}$ does not occur in $\gvConf{G}$, $\ren{\gvConf{C}}(\gvVar{c})$ occurs at most once in $\ftm{\gvConf{G}}$, \ie $\ftmArc{c}{\gvConf{C}}\le{1}$. Apply \textsc{E-Zap}.
  \item
    Case \textsc{E-Raise}.
    We have \\
    $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvPlug{\gvConf{F}}{\gvTry{\gvPlug{P}{\gvRaise}}{x}{M}{N}}}$. \\
    Apply \cref{lem:translate-raise} and \textsc{E-None}. \\
    For each channel $\gvVar{c}$ in $\gvVar{P}$: As $\tm{\gvConf{C}}$ is well-typed, $\gvVar{c}$ does not occur in $\gvConf{G}$, and $\ren{\gvConf{C}}(\gvVar{c})$ occurs at most once in $\ftm{\gvConf{G}}$, \ie $\ftmArc{c}{\gvConf{C}}\le{1}$. Apply \textsc{E-Zap}.
  \item
    Case \textsc{E-RaiseChild}.
    We have
    $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvChild{\gvPlug{P}{\gvRaise}}}$. \\
    Apply \cref{lem:translate-raise} and \textsc{E-HaltChild}.
  \item
    Case \textsc{E-RaiseMain}.
    We have
    $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvMain{\gvPlug{P}{\gvRaise}}}$. \\
    Apply \cref{lem:translate-raise} and \textsc{E-HaltMain}.
  \item
    Case \textsc{E-GC}.
    We have \\
    $\tm{\gvConf{C}} = \gvPlug{\gvConf{G}}{\gvPar{\gvRes{a}{\gvRes{b}{\gvPPar{\gvZap{a}}{\gvPar{\gvZap{b}}{\gvBuf{a}{\gvVecEmp}{b}{\gvVecEmp}}}}}}{\gvConf{C}}}$. \\
    As $\tm{\gvConf{C}}$ is well-typed, $\gvVar{a}$ and $\gvVar{b}$ do not occur in $\gvConf{G}$, and $\ren{\gvConf{C}}(\gvVar{a})$ does not occur in $\ftm{\gvConf{G}}$, \ie $\ftmArc{a}{\gvConf{C}}=0$. \\
    Apply \textsc{E-GC}.
  \item
    Case \textsc{E-LiftM}.
    Apply \cref{lem:translate-term} and \textsc{E-LiftM}.
  \end{itemize}
  In the remaining cases, apply the rule of the same name.
\end{proof}

\begin{theorem}
  If $\gvConf{C}$ is well-typed and $\tm{\ftm{\gvConf{C}}}\gvRedArr\tm{\gvConf{D}}$, then there exists a term $\tm{\gvConf{C'}}$ such that $\tm{\gvConf{C}}\equiv\gvRedArr^{+}\tm{\gvConf{C'}}$ and $\tm{\gvConf{D}}\gvRedArr^{\star}\tm{\ftm{\gvConf{C'}}}$.
\end{theorem}
\begin{proof}
  By case analysis on the reduction.
  \begin{itemize}
  \item
    Case \textsc{E-Zap}. There must be a channel $\gvVar{a}$ for which $\ftmArc{a}{\gvConf{C}}\le{1}$. There are two cases:
    \begin{itemize}
    \item 
      Case $\gvConf{C}\equiv\gvPlug{\gvConf{G}}{\gvPlug{\gvConf{F}}{\gvCancel{a}}}$. Let $\gvConf{C'}=\gvPlug{\gvConf{G}}{\gvPar{\gvPlug{\gvConf{F}}{\gvUnit}}{\gvZap{a}}}$. \\ For $\gvConf{C}\equiv\gvRedArr\gvConf{C'}$: Apply \textsc{E-Cancel}. \\ For $\rvConf{D}\rvRedArr^{\star}\ftm{\gvConf{C'}}$: Apply reflexivity.
    \item
      Case $\gvConf{C}\equiv\gvPlug{\gvConf{G}}{\gvPar{\gvZap{a}}{\gvBuf{a}{\gvVar{U}\gvVecAdd\gvVec{V}}{b}{\gvVec{W}}}}$. \\ Let $\gvConf{C'} = \gvPlug{\gvConf{G}}{\gvPar{\gvZap{a}}{\gvPar{\gvZap{U}}{\gvBuf{a}{\gvVec{V}}{b}{\gvVec{W}}}}}$. \\ For $\gvConf{C}\equiv\gvRedArr\gvConf{C'}$: Apply \textsc{E-Zap}. \\ For $\rvConf{D}\rvRedArr^{\star}\ftm{\gvConf{C'}}$: For each channel $\gvVar{c}$ in $\gvVar{U}$: $\tm{\gvConf{C'}}$ is well-typed, by subject reduction, and hence $\gvVar{c}$ does not occur in $\gvConf{G}$, and $\ren{\gvConf{C'}}(\gvVar{c})$ occurs at most once in $\ftm{\gvConf{G}}$, \ie $\ftmArc{c}{\gvConf{C}}\le{1}$. Apply \textsc{E-Zap}.
    \end{itemize}
  \item
    Case \textsc{E-GC}. We have $\ftmArc{a}{\gvConf{C}}=0$. Therefore, there must be two channels $\tm{a}$ and $\tm{b}$ which are cancelled, \ie $\gvConf{C}\equiv\gvPlug{\gvConf{G}}{\gvRes{a}{\gvRes{b}{\gvPar{\gvPPar{\gvZap{a}}{\gvPar{\gvZap{b}}{\gvBuf{a}{\gvVec{V}}{b}{\gvVec{W}}}}}{\gvConf{E}}}}}$. \\ Let $\gvConf{C'}=\gvPlug{\gvConf{G}}{\gvConf{E}}$. \\
    For $\gvConf{C}\equiv\gvRedArr\gvConf{C'}$: Apply \textsc{E-Zap} repeatedly. Apply \textsc{E-GC}. \\
    For $\rvConf{D}\gvRedArr^{\star}\ftm{\gvConf{C'}}$: Apply reflexivity.
  \item
    Case \textsc{E-HaltChild}.
    There are two cases:
    \begin{itemize}
    \item
      Case $\gvConf{C}\equiv\gvPlug{\gvConf{G}}{\gvPar{\gvConf{E}}{\gvChild{\gvUnit}}}$. Let $\gvConf{C'}=\gvPlug{\gvConf{G}}{\gvConf{E}}$. \\ For $\gvConf{C}\equiv\gvRedArr\gvConf{C'}$: Apply \textsc{E-HaltChild}. \\ For $\rvConf{D}\rvRedArr\ftm{\gvConf{C'}}$: Apply reflexivity.
    \item
      Case $\gvConf{C}\equiv\gvPlug{\gvConf{G}}{\gvPar{\gvConf{E}}{\gvChild{\gvRaise}}}$. Let $\gvConf{C'}=\gvPlug{\gvConf{G}}{\gvConf{E}}$. \\ For $\gvConf{C}\equiv\gvRedArr\gvConf{C'}$: Apply \textsc{E-RaiseChild}. \\ For $\rvConf{D}\rvRedArr\ftm{\gvConf{C'}}$: Apply reflexivity.
    \end{itemize}
  \end{itemize}
\end{proof}

### 2019-05-05 src/lib.rs

@@ -141,7 +141,6 @@ pub type Offer<S1, S2> =
 pub type Choose<S1, S2> =
     Send<Either<<S1 as Session>::Dual, <S2 as Session>::Dual>, End>;
 
-
 pub fn offer_either<'a, S1, S2, F, G, R>(s: Offer<S1, S2>, f: F, g: G)
                                          -> Result<R, Box<Error + 'a>>
 where
@@ -151,30 +150,30 @@ where
     G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
 {
     let (e, s) = recv(s)?;
-    close(s)?;
+    cancel(s);
     e.either(f, g)
 }
 
-pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> Result<S1, Box<Error>>
+pub fn choose_left<'a, S1, S2>(s: Choose<S1, S2>) -> S1
 where
     S1: Session + 'a,
     S2: Session + 'a,
 {
     let (here, there) = S1::new();
     let s = send(Either::Left(there), s);
-    close(s)?;
-    Ok(here)
+    cancel(s);
+    here
 }
 
-pub fn choose_right<'a, S1, S2>(s: Choose<S1, S2>) -> Result<S2, Box<Error>>
+pub fn choose_right<'a, S1, S2>(s: Choose<S1, S2>) -> S2
 where
     S1: Session + 'a,
     S2: Session + 'a,
 {
     let (here, there) = S2::new();
     let s = send(Either::Right(there), s);
-    close(s)?;
-    Ok(here)
+    cancel(s);
+    here
 }
 
 
@@ -185,7 +184,7 @@ macro_rules! offer {
     ($session:expr, { $($pat:pat => $result:expr,)* }) => {
         (move || -> Result<_, _> {
             let (l, s) = recv($session)?;
-            close(s)?;
+            cancel(s);
             match l {
                 $(
                     $pat => $result,
@@ -197,16 +196,12 @@ macro_rules! offer {
 
 #[macro_export]
 macro_rules! choose {
-    ($label:path, $session:expr) => {
-        (move || -> Result<_, _> {
-            let (here, there) = <_ as Session>::new();
-            let s = send($label(there), $session);
-            match close(s) {
-                Ok(()) => Ok(here),
-                Err(e) => Err(e),
-            }
-        })()
-    };
+    ($label:path, $session:expr) => {{
+        let (here, there) = <_ as Session>::new();
[…]

### 2019-05-05 tests/benchmark_tests.rs

@@ -60,11 +60,11 @@ mod rusty_variation_bench {
 
     fn neg_client(s: CalcCli) -> Result<(), Box<Error>> {
         let n = thread_rng().gen();
-        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s)?;
+        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s);
         let s = send(n, s);
         let (n_, s) = recv(s)?;
         assert_eq!(-n, n_);
-        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s)?;
+        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s);
         close(s)?;
         Ok(())
     }

### 2019-05-05 tests/tests.rs

@@ -67,7 +67,7 @@ fn simple_calc_works() {
         {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
-            let s = choose_left::<_, AddClient<i32>>(s)?;
+            let s = choose_left::<_, AddClient<i32>>(s);
             let s = send(x, s);
             let (y, s) = recv(s)?;
             close(s)?;
@@ -79,7 +79,7 @@ fn simple_calc_works() {
             let s: SimpleCalcClient<i32> = fork(simple_calc_server);
             let x: i32 = rng.gen();
             let y: i32 = rng.gen();
-            let s = choose_right::<NegClient<i32>, _>(s)?;
+            let s = choose_right::<NegClient<i32>, _>(s);
             let s = send(x, s);
             let s = send(y, s);
             let (z, s) = recv(s)?;
@@ -131,7 +131,7 @@ fn nice_calc_works() {
         {
             let s: NiceCalcClient<i32> = fork(nice_calc_server);
             let x: i32 = rng.gen();
-            let s = choose!(CalcOp::Neg, s)?;
+            let s = choose!(CalcOp::Neg, s);
             let s = send(x, s);
             let (y, s) = recv(s)?;
             close(s)?;
@@ -143,7 +143,7 @@ fn nice_calc_works() {
             let s: NiceCalcClient<i32> = fork(nice_calc_server);
             let x: i32 = rng.gen();
             let y: i32 = rng.gen();
-            let s = choose!(CalcOp::Add, s)?;
+            let s = choose!(CalcOp::Add, s);
             let s = send(x, s);
             let s = send(y, s);
             let (z, s) = recv(s)?;
@@ -228,7 +228,7 @@ fn closure_works() {
 
         // Create a closure which uses the session.
         let f = move |x: i32| -> Result<i32, Box<Error>> {
-            let s = choose!(CalcOp::Neg, s)?;
+            let s = choose!(CalcOp::Neg, s);
             let s = send(x, s);
             let (y, s) = recv(s)?;
             close(s)?;
@@ -277,12 +277,12 @@ fn nice_sum_client_accum(s: NiceSumClient<i32>, mut xs: Vec<i32>)
                          -> Result<i32, Box<Error>> {
     match xs.pop() {
         Option::Some(x) => {
-            let s = choose!(SumOp::More, s)?;
+            let s = choose!(SumOp::More, s);
             let s = send(x, s);
[…]

### 2019-05-20 src/lib.rs
Added homogeneous select operator

@@ -7,7 +7,7 @@ use std::marker;
 use std::mem;
 use std::thread::{JoinHandle, spawn};
 use std::panic;
-use crossbeam_channel::{Sender, Receiver, bounded};
+use crossbeam_channel::{Sender, Receiver, bounded, Select};
 use either::Either;
 
 /// The session types supported.
@@ -71,17 +71,17 @@ impl<T: marker::Send, S: Session> Session for Recv<T, S> {
 
 // Send and receive
 
-pub fn send<'a, T, S>(x: T, s: Send<T, S>) -> S
+pub fn send<T, S>(x: T, s: Send<T, S>) -> S
 where
-    T: marker::Send + 'a,
-    S: Session + 'a,
+    T: marker::Send,
+    S: Session,
 {
     let (here, there) = S::new();
     s.channel.send((x, there)).unwrap_or(());
     here
 }
 
-pub fn recv<'a, T, S>(s: Recv<T, S>) -> Result<(T, S), Box<Error + 'a>>
+pub fn recv<T, S>(s: Recv<T, S>) -> Result<(T, S), Box<Error>>
 where
     T: marker::Send,
     S: Session,
@@ -207,19 +207,44 @@ macro_rules! choose {
 
 // Selection
 
-// pub fn select<'a, T, S, I>(rs: &[Recv<T, S>]) -> Result<(T, S), Box<Error + 'a>>
-// where
-//     T: marker::Send + 'a,
-//     S: Session + 'a,
-// {
-//     let mut sel = Select::new();
-//     for r in rs {
-//         sel.recv(&r.channel);
-//     }
-//     loop {
-//         let oper = sel.select();
-//         let index = oper.index();
-//         let (x, s) = oper.recv(&rs[index].channel)?;
-//         break Ok((x, s));
-//     }
-// }
+pub fn select_mut<T, S, I>(rs: &mut Vec<Recv<T, S>>) -> Result<(T, S), Box<Error>>
+where
+    T: marker::Send,
+    S: Session,
+{
+    let (index, res) = {
+        let mut sel = Select::new();
+        let iter = rs.iter();
+        for r in iter {
+            sel.recv(&r.channel);
+        }
+        loop {
+            let index = sel.ready();
+            let res = rs[index].channel.try_recv();
+
+            if let Err(e) = res {
+                if e.is_empty() {
+                    continue;
+                }
+            }
+
[…]

### 2019-05-20 src/lib.rs
Wrote a test for selection

@@ -3,6 +3,7 @@ extern crate either;
 
 use std::boxed::Box;
 use std::error::Error;
+use std::fmt;
 use std::marker;
 use std::mem;
 use std::thread::{JoinHandle, spawn};
@@ -207,44 +208,77 @@ macro_rules! choose {
 
 // Selection
 
-pub fn select_mut<T, S, I>(rs: &mut Vec<Recv<T, S>>) -> Result<(T, S), Box<Error>>
+#[derive(Debug)]
+enum SelectError {
+    EmptyVec,
+}
+
+impl fmt::Display for SelectError {
+    fn fmt(&self, f: &mut fmt::Formatter) -> fmt::Result {
+        match *self {
+            SelectError::EmptyVec =>
+                write!(f, "please use a vector with at least one element"),
+        }
+    }
+}
+
+impl Error for SelectError {
+    fn description(&self) -> &str {
+        match *self {
+            SelectError::EmptyVec => "empty vectors not allowed",
+        }
+    }
+
+    fn cause(&self) -> Option<&Error> {
+        match *self {
+            SelectError::EmptyVec => None,
+        }
+    }
+}
+
+pub fn select_mut<T, S>(rs: &mut Vec<Recv<T, S>>) -> Result<(T, S), Box<Error>>
 where
     T: marker::Send,
     S: Session,
 {
-    let (index, res) = {
-        let mut sel = Select::new();
-        let iter = rs.iter();
-        for r in iter {
-            sel.recv(&r.channel);
-        }
-        loop {
-            let index = sel.ready();
-            let res = rs[index].channel.try_recv();
-
-            if let Err(e) = res {
-                if e.is_empty() {
-                    continue;
-                }
+    if rs.is_empty() {
+        Err(Box::new(SelectError::EmptyVec))
+    }
+    else {
+        let (index, res) = {
+            let mut sel = Select::new();
+            let iter = rs.iter();
+            for r in iter {
+                sel.recv(&r.channel);
             }
+            loop {
+                let index = sel.ready();
+                let res = rs[index].channel.try_recv();
 
-            break (index, res);
+                if let Err(e) = res {
+                    if e.is_empty() {
+                        continue;
+                    }
+                }
+
+                break (index, res);
+            }
+        };
+
[…]

### 2019-05-20 tests/tests.rs

@@ -8,6 +8,8 @@ use std::error::Error;
 use std::marker;
 use std::mem;
 use std::sync::mpsc;
+use std::thread::sleep;
+use std::time::Duration;
 
 
 // Test sending a ping across threads.
@@ -309,6 +311,44 @@ fn recursion_works() {
     }().is_ok());
 
     assert!(other_thread.join().is_ok());
+}
+
+// Test selection.
+
+#[test]
+fn selection_works() {
+
+    let mut other_threads = Vec::new();
+    let mut rs = Vec::new();
+
+    for i in 0..10 {
+        let (other_thread, s) = fork_with_thread_id(move |s: Send<u64, End>| {
+            sleep(Duration::from_millis(i * 10));
+            let s = send(i, s);
+            close(s)
+        });
+        other_threads.push(other_thread);
+        rs.push(s);
+    }
+
+    assert!(|| -> Result<(), Box<Error>> {
+        let mut current_index = 0;
+        loop {
+            if rs.is_empty() {
+                break Ok(());
+            }
+            else {
+                let (i, r) = select_mut(&mut rs)?;
+                close(r)?;
+                assert_eq!(current_index, i);
+                current_index += 1;
+            }
+        }
+    }().is_ok());
+
+    for other_thread in other_threads {
+        assert!(other_thread.join().is_ok())
+    }
 }
 
 

### 2019-06-28 doc/ice2019/main.tex
Added ICE slides; renamed abstract to ICE paper; wrote on paper; added doctest for session reuse.

\documentclass[copyright,creativecommons]{eptcs}
\providecommand{\event}{ICE 2019} % Name of the event you are submitting to
\input{preamble}

\title{Rusty Variation\\ Deadlock-free Sessions with Failure in Rust}
\author{Wen Kokke
  \institute{LFCS, University of Edinburgh, 10 Crichton St, EH8 9AB, Edinburgh, United Kingdom}
  \email{wen.kokke@ed.ac.uk}
}
\def\titlerunning{Rusty Variation}
\def\authorrunning{Wen Kokke}

\begin{document}

\maketitle

\begin{abstract}
  Rusty Variation (RV) is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees. Programs written using RV are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
\end{abstract}

\section{Introduction}\label{sec:introduction}
In concurrent programming, processes commonly exchange information by sending messages on shared channels. However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.

Session types are a substructural typing discipline, introduced by Honda~\cite{honda1993}, which describe communication protocols as \emph{channel types}, capturing both the type and the order of messages. Session type systems, then, ensure that programs correctly implement specified protocols.

Rust is an immensely popular systems programming language.\footnote{\url{https://insights.stackoverflow.com/survey/2019/}} It aims to combine efficiency with abstraction and thread and memory safety. However, out of the box, it does not offer a way to specify protocols and verify communication.

In this paper, we present Rusty Variation\footnote{\url{https://github.com/wenkokke/rusty-variation/}} (RV), a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure. It is these strong guarantees of correctness, together with out account of failure, which sets us apart from previous work~\cite{jespersen2015}. We assume some familiarity with Rust. For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}.\footnote{\url{https://doc.rust-lang.org/nightly/book/}}

We base our work on Exceptional GV by Fowler \etal~\cite[EGV]{fowler2019}. Exceptional GV is a descendant of Good Variation~\cite[GV]{wadler2012} which adds support for exceptions. Crucially, EGV has an affine type system, and permits the explicit cancellation of sessions. EGV guarantees session fidelity, global progress, freedom from deadlocks and livelocks, confluence, and termination. That is to say, there are no locks or race conditions, and all programs eventually halt. For a discussion of EGV and its metatheory, we refer the reader to the work by Fowler \etal~\cite{fowler2019}.

Rusty Variation is an implementation of Exceptional GV (EGV) in Rust. We claim that RV is a faithful implementation of EGV, preserving much of EGV's metatheory, meaning RV guarantees session fidelity, global progress, confluence, and freedom from deadlocks and unwanted races. However, we depart from EGV in two ways:
First, EGV relies upon exceptions and handlers. Rust, however, rejects exceptions in favour of monadic errors, \ie, wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.
Second, The core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee livelock freedom, or---well---termination.

The paper proceeds as follows. In~\cref{sec:contribution}, we describe the programming interface offered by Rusty Variation, and provide several example programs. In~\cref{sec:related-work}, we compare Rusty Variation to the related work by Laumann \etal~\cite{jespersen2015}. Finally, in~\cref{sec:future-work}, we describe future work on verifying the correctness guarantees offered by RV.

\section{Contribution: Rusty Variation}\label{sec:contribution}

\subsection{Types, sessions, and duality}
There are three basic types which we use to describe session protocols---or \emph{session types}: \lstinline{Send}, \lstinline{Recv}, and \lstinline{End}. We refer to values of these types as \emph{session endpoints} or, when referring to both endpoints, \emph{sessions}. Session endpoints wrap primitive channels---see~\cref{sec:implementation}. Any type built using these primitive types is a valid session type:
\begin{lstlisting}
pub struct Send<T, S: Session> { ... }
pub struct Recv<T, S: Session> { ... }
pub struct End { ... }
\end{lstlisting}
Session types are public, but their member functions are not. Therefore, users cannot construct their own misbehaving sessions. The continuation of a session must itself be a session. This is enforced by the \lstinline{Session} constraint. In addition to this ``kinding'' of session types, the \lstinline{Session} trait also requires any session type to implement duality and session generation:
\begin{lstlisting}
pub trait Session {
  type Dual: Session<Dual=Self>;
  fn new() -> (Self, Self::Dual);
}
\end{lstlisting}
Duality works as expected: \lstinline{Send<T, S>} is dual to \lstinline{Recv<T, S::Dual>}---and vice versa---and \lstinline{End} is self-dual. The constraint \lstinline{Session<Dual=Self>} enforces that the dual of a session type is a session type, and that \lstinline{S::Dual::Dual} is equal to \lstinline{S} (involutivity).

The \lstinline{new} function generates a new session-typed channel and returns two \emph{dual} channel endpoints. Unlike the rest of the functions discussed in this section, \lstinline{new} is \emph{not} part of the public interface---see \cref{sec:limitations}.

Using these types, we can define the types of servers which offer squaring or negation as follows, where \lstinline{i32} is the type of 32-bit integers:
\begin{lstlisting}
type SqrSrv = Recv<i32, Send<i32, End>>;
type NegSrv = Recv<i32, Send<i32, End>>;
\end{lstlisting}
The client types are obtained by duality.

\subsection{Send, receive, close, and fork}\label{sec:basics}
The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. Sending is non-blocking, and never fails. Sending a value on a cancelled channel is equivalent to cancelling that value---see \cref{sec:cancellation}. Receiving is blocking, and fails if the channel is cancelled.
\begin{lstlisting}
pub fn send<T, S: Session>(x: T, s: Send<T, S>) -> S
pub fn recv<T, S: Session>(s: Recv<T, S) -> Option<(T, S)>
\end{lstlisting}
The \lstinline{close} function closes a \emph{completed} session, \ie of type \lstinline{End}. Closing is \emph{synchronous}, meaning that \lstinline{close} blocks until the dual end point is also closed. Closing a cancelled channel fails.
\begin{lstlisting}
pub fn close(s: End) -> Option<()>
\end{lstlisting}
The \lstinline{fork} function creates a new session, and spawns a new thread. The new thread runs the first argument of \lstinline{fork}---a function---which receives one endpoint, and \lstinline{fork} returns the dual endpoint. It has the following type---where \lstinline{FnOnce(S) -> Option<()>} is the type of functions from \lstinline{S} to \lstinline{Option<()>} which can be invoked only once:
\begin{lstlisting}
pub fn fork<S: Session>(p: P) -> S::Dual
where
  P: FnOnce(S) -> Option<()>,
\end{lstlisting}
The function passed to \lstinline{fork} is allowed to fail, hence the \lstinline{Option}-type. If it fails, the child thread panics silently, and all open sessions in the child thread are cancelled.

Now that we have covered the basics, let us consider an example program. The following program forks off a child process which sends a ``ping'', then waits to receive that ping and returns:
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  let s = send((), s);
  close(s)
});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
The \lstinline{let x = M?; N} construct is Rust's ``monadic bind'' notation for programs which may return errors. If \lstinline{recv(s)} succeeds, the \lstinline{?}-operator unpacks the \lstinline{Option}. If \lstinline{recv(s)} fails, the \lstinline{?}-operator short-circuits, skips the rest of the statements, and returns the error.

\subsection{Linearity and cancellation}\label{sec:cancellation}
Session types often require linearity---the protocol for a negation server in~\cref{sec:basics} specifies that the server receives \emph{one} integer, and then sends \emph{one} integer back. For such a server to, \eg send back multiple integers would be, at best, confusing, and at worst, cause an error. The following two erroneous programs violate the linearity condition. The program on the left reuses the channel s, and the program on the right drops it without completing the session:

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  let s1 = send((), s);
  close(s1)?;
  let s2 = send((), ~s~); // reuse s
  close(s2)
});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {

  // unintentionally
  // left blank.

});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
\end{minipage}

\noindent
Rust is an affine language, meaning that any value can be used \emph{at most once}. Hence, our first example doesn't compile: Rust complains about the second usage of \lstinline{s}, which attempts to use \lstinline{s} after transferring ownership of the value to \lstinline{send}.

The second program is trickier, as Rust's type system doesn't help us here. Arguably, however, that is for the best. The idea of \emph{linear} session types ignores the reality that sessions may be dropped at any time, for any number of reasons: a process may panic, dropping all sessions in scope, or---when communicating over the network---the connection might fail. In~\cref{sec:introduction}, we mentioned that EGV has an affine type system, and permits the explicit cancellation of sessions. Following EGV, Rusty Variation also permits the cancellation of sessions. However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/nomicon/destructors.html}}.

When an initialised value in Rust goes out of scope, Rust calls its destructor, the \lstinline{drop} method. This method can be customised by implementing the \lstinline{Drop} trait. For instance, if we use the implementation of \lstinline{Drop} shown on the left, running the code on the right would print ``End Dropped!'' twice, once from the child thread, and once from the main thread.

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
impl Drop for End {
  fn drop(&mut self) {
    println!("End Dropped!")
  }
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let _s = fork(move |_s: End| {
  // intentionally left blank.
});
// intentionally left blank.
\end{lstlisting}
\end{minipage}

\noindent
To mimic the cancellation behaviour in EGV, we count the number of references to a session, starting at \emph{two} when a session is created using \lstinline{new}. Each time a session's destructor is called, we decrement this counter. If the counter is \emph{one}, the session is marked as \emph{disconnected}. Sending on a disconnected session drops the value being sent. Receiving on or closing disconnected sessions returns an error, \ie \lstinline{None}. This ensures that no process is left waiting for a message that will never come. If the counter reaches \emph{zero}, the memory for the session is deallocated.

As it turns out, this is exactly the behaviour for the channels from \lstinline{crossbeam_channel}, which we use in our implementation, and when a session is dropped, the channel it wraps is dropped recusively. Hence, we inherit the correct behaviour of dropping sessions from \lstinline{crossbeam_channel}.

However, there is one major downside to replacing explicit cancellation with implicit cancellation: it is no longer possible to tell whether the programmer wanted to cancel a session or simply forgot to complete the session. We do two things to remedy this. First, the definitions of \lstinline{Send}, \lstinline{Recv}, and \lstinline{End} are annotated with \lstinline{#[must_use]}. This causes Rust to emit a warning whenever a session is dropped. Second, we provide an explicit \lstinline{cancel} function, though its implementation is simply to drop its argument:
\begin{lstlisting}
  pub fn cancel<T>(_x: T) -> () { /* intentionally left blank. */ }
\end{lstlisting}

The following two programs illustrate our solution. The program on the left forgets to use \lstinline{s}, and Rust emits a warning, complaining that \lstinline{s} isn't used. The program on the right passes the session to \lstinline{cancel}, explicitly marking the drop, and hence compiles without any warnings. Both programs have the same semantics, though: the program forks off a process which is \emph{expected} to send a ``ping''. Instead the forked process cancels the session---be it implicitly or explicitly. Hence, the call to \lstinline{recv} fails, the \lstinline{?}-operator short-circuits, and the whole program returns \lstinline{None}:

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  // shouldn't we be using s?
  Some(())
});
let (z, s) = recv(s)?;
close(s)
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let s = fork(move |s: Send<(), End>| {
  cancel(s);
  Some(())
});
let (z, s) = recv(s)?;
close(s)
\end{lstlisting}
\end{minipage}%

\subsection{Branching with offer! and choose!}\label{sec:choice}
The \lstinline{choose_left}, \lstinline{choose_right}, and \lstinline{offer_either} functions let processes offer or make a \emph{binary} choice. These are \emph{derived} constructs~\cite{honda1993,dardha2017}---they are implemented using only the other functions in our library. They send or receive a value of the sum type, \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.
\begin{lstlisting}
type Choose<S1, S2> = Send<Either<S1, S2>, End>;
type Offer<S1, S2> = Recv<Either<S1, S2>, End>;

pub fn choose_left<S1: Session, S2: Session>(s: Choose<S1, S2>) -> S1
pub fn choose_right<S1: Session, S2: Session>(s: Choose<S1, S2>) -> S2
pub fn offer_either<S1: Session, S2: Session, P1, P2, R>(
       s: Offer<S1, S2>, p1: P1, p2: P2) -> Option<R>
where
  P1: FnOnce(S1) -> Option<R>,
  P2: FnOnce(S2) -> Option<R>,
\end{lstlisting}
There is a subtlety in their implementation. The \lstinline{choose_left} function creates a new session of type \lstinline{S1}. It wraps one endpoint in \lstinline{Either::Left}, sends it over the original session \lstinline{s}, and returns the other endpoint. However, this leaves it with the continuation of the original session, of type \lstinline{End}. Closing this endpoint synchronously, using \lstinline{close}, may fail. However, \lstinline{choose_left} \emph{sends}, and for consistency with \lstinline{send}, we would like it to always succeed. Therefore, we opt to \emph{cancel} the continuation instead:
\begin{lstlisting}
pub fn choose_left<S1: Session, S2: Session>(s: Choose<S1, S2>) -> S1 {
    let (here, there) = S1::new();
    let s = send(Either::Left(there), s);
    cancel(s);
    here
} 
\end{lstlisting}
Cancelling the endpoint is harmless---as long as the other processes is prepared, and uses the corresponding function \lstinline{offer_either}, this acts as an \emph{asynchronous} alternative to close. Pairing \lstinline{choose_left} with a manually implemented version of \lstinline{offer_either} which attempts to \emph{close} the continuation instead of cancelling it results in a runtime error.

The \lstinline{choose_left}, \lstinline{choose_right}, and \lstinline{offer_either} functions suffice to encode choice, they are not useful in practice---a protocol with a four-way choice would require three nested \lstinline{Offer} constructs, and several calls to \lstinline{offer_either}. Code written in this style rapidly becomes unreadable. For this reason, we also define two macros, \lstinline{choose!} and \lstinline{offer!}, which generalise binary choice using \lstinline{Either} to a choice on \emph{any} \lstinline{enum}, as long as each case wraps a session. This allows processes to offer and make choices between many \emph{labelled} branches, and allows us to define the type for a calculator server as:
\begin{lstlisting}
enum CalcOp { Sqr(SqrSrv), Neg(NegSrv) }
type CalcSrv = Recv<CalcOp, End>;
type CalcCli = CalcSrv::Dual;
\end{lstlisting}
Using the \lstinline{choose!} and \lstinline{offer!} macros, we can implement a calculator server and client. The \lstinline{main} function below forks off a calculator server, which offers the choice between squaring and negation. The client then selects squaring, sends the number four, and prints the result:

\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
fn server(s: CalcSrv) -> Option<()> {
  offer!(s, {
    CalcOp::Sqr(s) => {
      let (x, s) = recv(s)?;
      let s = send(x * x, s);
      close(s)
    },
    CalcOp::Neg(s) => {
      let (x, s) = recv(s)?;
      let s = send(-x, s);
      close(s)
    },
  })
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
fn client(s: CalcCli) -> Option<i32> {
  let s = choose!(CalcOp::Sqr, s);
  let s = send(4, s);
  let (z, s) = recv(s)?;
  close(s)?;
  Some(z)
}

// main() prints Some(16)
fn main() {
  let s = fork(server);
  let z = client(s);
  println!("{}", z);
}
\end{lstlisting}
\end{minipage}

\subsection{Selection}\label{sec:selection}
Selection is a mechanism which allows processes to block on a list of actions, waiting until the first of them fires. Writing programs which use selection can be tricky and error prone, even using libraries specifically built to help with selection. For instance, the following is an excerpt from the documentation of \lstinline{crossbeam_channel::Select}\footnote{\url{https://docs.rs/crossbeam-channel/0.1.2/crossbeam_channel/struct.Select.html}}, a construct which implements a selection mechanism:
\begin{quotation}
  \noindent
  Rules which must be respected in order for selection to work properly:
  \begin{enumerate}
  \item Before each selection, a fresh Select must be created.
  \item Selection cases must be repeatedly probed in a loop.
  \item If a selection case fires, the loop must be broken without probing cases any further.
  \item In each iteration of the loop, the same set of cases must be probed in the same order.
  \item No selection case may be repeated.
  \item No two cases may operate on the same end (receiving or sending) of the same channel.
  \item There must be at least one send, or at least one recv case.
  \end{enumerate}
\end{quotation}
None of these rules are enforced by Rust---instead, all of them must be checked by the programmer---and bugs in implementing selection can be hard to find.

RV offers an alternative, the \lstinline{select_mut} function. This function operates on a vector of \emph{receiving} endpoints, returning the result of the first receive operation which completes, and removing the corresponding session from the vector. We restrict ourselves to the receiving endpoints, as RV's \lstinline{send} operation is non-blocking, hence selecting on sending endpoints doesn't make any sense.
\begin{lstlisting}
pub fn select_mut<T, S: Session>(rs: &mut Vec<Recv<T, S>>) -> Option<(T, S)>
\end{lstlisting}

The \lstinline{select_mut} function is implemented on top of \lstinline{crossbeam_channel::Select}, in a way which respects the rules listed above. In addition, RV offers an immutable variant of \lstinline{select_mut}, named \lstinline{select}, which returns the remainder of the endpoints along the with result of the selection. Using \lstinline{select_mut}, we can write the following program, which forks off ten threads, each sending an index between 1 and 10, and then repeatedly selects on the receiving endpoints, receiving and printing an index, printing the numbers 1 to 10 in a random order:

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
let mut rs = Vec::new();
for i in 0..10 {
  let s = fork(move |s: Send<i32, End>| {
    let s = send(i, s);
    close(s)
  });
  rs.push(s);
}
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
loop {
  if rs.is_empty() {
    break Ok(());
  }
  let (i, r) = select_mut(&mut rs)?;
  println!("{}", i);
  close(r)
}
\end{lstlisting}
\end{minipage}

The \lstinline{select_mut} and \lstinline{select} functions are homogeneous: they only select over vectors of endpoints with \emph{the same type}. However, this does not rule out heterogeneous uses of selection, where the operations are of different types. It is possible to perform heterogeneous selection by defining an enumeration, in the style of the \lstinline{CalcOp} type mentioned in~\cref{sec:choice}.

TODO: THIS IS WHERE I STOPPED

\subsection{Implementation}\label{sec:implementation}
We use the asynchronous channels from \lstinline{crossbeam_channel}. These are ideal, as they already implement the correct cancellation behaviour. Channels are typed, and can only transfer values of a single type. Therefore, we use one-shot channels, and encode sessions following Scalas and Yoshida, and Padovani~\cite{scalas2016,padovani2017}: each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value, and each \lstinline{recv} receives a value together with a channel on which to continue the session. Fortunately, the channels in \lstinline{crossbeam_channel} are highly optimised for one-shot usage.

\subsection{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Rusty Variation lock. The first is by looping, which we discussed earlier. The other two are explicit, and surprisingly, \lstinline{unsafe} is not one of them.

The biggest issue lies in the fact that Rust's semantics do not guarantee that destructors are actually run. This is important for manually managed memory, \eg doubly linked lists. For this reason, it is always possible to ``forget'' a value, using \lstinline{std::mem::forget}, without running its destructors.

Secondly, it is not possible---at the time of writing---to export a trait without exporting all its members. Therefore, any user can use the \lstinline{new} function, and doing so, it is trivial to write a program which deadlocks.

\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Deadlock using std::mem::forget
let s = fork(move |s: Send<(), End>| {
  std::mem::forget(s);
  Some(())
});
let ((), s) = recv(s)?;
close(s)
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Deadlock using new and spawn
let (s1, r1) = Send<Void, End>::new();
let (s2, r2) = Send<Void, End>::new();
std::thread::spawn(move || {
  let (v, r1) = recv(r1)?;
  close(r1)?;
  let s2 = send(v, s2);
  close(s2)
});
let (v, r2) = recv(r2)?;
close(r2)?;
let s1 = send(v, s1);
close(s1)
\end{lstlisting}
\end{minipage}

\section{Related Work}\label{sec:related-work}
Laumann \etal~\cite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by Gay and Vasconcelos~\cite{gay2009}. Their implementation guarantees that programs correctly implement specified communication protocols with some caveats. LAST depends upon linear types, meaning that sessions have to be used exactly once. Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped. The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}. In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \lstinline{#[must_use]} annotation. However, in general, this is not something the compiler can guarantee. Furthermore, linear session types ignore the reality that channels may be dropped at any time due to, \eg a thread panicking.

We embrace the affine, and base our work on Exceptional GV by Fowler \etal~\cite[EGV]{fowler2019}. Exceptional GV is a descendant of Good Variation~\cite[GV]{wadler2012}, which itself is a descendant of LAST. Crucially, EGV has an affine type system, and allows for the explicit cancellation of sessions. Furthermore, EGV offers a stronger metatheory than LAST. In addition to session fidelity, EGV also guarantees global progress, freedom from deadlocks and livelocks, confluence, and termination. That is to say, there are no locks or race conditions, and all programs eventually halt. For a discussion of EGV and its metatheory, we refer the reader to the work by Fowler \etal~\cite{fowler2019}.

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Rusty Variation
// main() returns None
fn server(_s: Recv<(), End>) {
  // unintentionally left blank.
}
fn client(s: Send<(), End>) {
  close(send((), s))
}
fn main() -> Option<()> {
  client(fork(server))
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
// main() causes segfault
fn server(_c: Chan<(), Recv<(), Eps>>) {
  // unintentionally left blank.
}
fn client(c: Chan<(), Send<(), Eps>>) {
  c.send(()).close();
}
fn main() -> () {
  connect(server, client);
}
\end{lstlisting}
\end{minipage}%

\vspace{-0.5\baselineskip}
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}
// Laumann et al.
// The order in the type doesn't match
// the order in the code. This compiles,
// not because the cases are associated
// correctly, but because the labels are
// not checked.

type SqrSrv  = Recv<i32, Send<i32, Eps>>;
type NegSrv  = Recv<i32, Send<i32, Eps>>;
type CalcSrv = Offer<SqrSrv, NegSrv>;

fn server(c: Chan<(), CalcSrv>) {
  offer!{c,
    NEG => {
      let (c, n) = c.recv();
      let c = c.send(-n);
      c.close();
    },
    SQR => {
      let (c, n) = c.recv();
      let c = c.send(n * n);
      c.close();
    }
  }
}
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\begin{lstlisting}

// The client selects the first option,
// assuming it represents squaring...
// Unfortunately, it represents negation.

type SqrCli  = Send<i32, Recv<i32, Eps>>;
type NegCli  = Send<i32, Recv<i32, Eps>>;
type CalcCli = Choose<SqrCli, NegCli>;

fn client(c: Chan<(), CalcCli>) {
  let c = c.sel1(); // select SqrCli
  let c = c.send(42);
  let (c, n) = c.recv();
  println!("{}", n);
  c.close();
}

fn main() {
  connect(server, client); // prints -42
}
\end{lstlisting}
\end{minipage}

\subsection{Future Work}\label{sec:future-work}
In \cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.

Firstly, RV is unit tested. We test various use-cases from EGV, from basic operations and their interactions with cancellation, to delegation and recursive session types. This test suite has caught several errors in the implementation. However, unit tests are error prone, and it is hard (if not impossible) to obtain a set of tests which cover all use-cases of a library. Unfortunately, tools such as RustBelt~\cite{jung2017} and Rust Distilled~\cite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of RV.

In the absence of a mature formal semantics for Rust, we are designing a formal language, RV (or ``formal RV''), which matches, as close as possible, the semantics of our Rust library. It is a concurrent lambda calculus with heap-based shared-memory semantics. We plan to prove that formal RV guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom. Furthermore, we plan to show that there exists a translation from EGV to formal RV, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\cite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV. This would show that formal RV is a faithful implementation of EGV.

Lastly, we plan to ensure the correspondence between formal RV and the implementation of RV, using property-based testing. We plan to generate a series of random formal RV terms, using Neat~\cite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal RV.

\bibliographystyle{eptcs}
\bibliography{main}

\end{document}

### 2019-06-28 tests/benchmark_tests.rs

@@ -85,6 +85,46 @@ mod session_types_bench {
     use test::Bencher;
     use session_types::*;
 
+    mod secret {
+
+        use session_types::*;
+
+        type AddSrv  = Recv<i32, Recv<i32, Send<i32, Eps>>>;
+        type SubSrv  = Recv<i32, Recv<i32, Send<i32, Eps>>>;
+        type CalcSrv = Offer<AddSrv, SubSrv>;
+
+        fn server(c: Chan<(), CalcSrv>) {
+            offer!{c,
+                   SUB => {
+                       let (c, n) = c.recv();
+                       let (c, m) = c.recv();
+                       let c = c.send(n - m);
+                       c.close();
+                   },
+                   ADD => {
+                       let (c, n) = c.recv();
+                       let (c, m) = c.recv();
+                       let c = c.send(n + m);
+                       c.close();
+                   }
+            }
+        }
+
+        type AddCli  = Send<i32, Send<i32, Recv<i32, Eps>>>;
+        type SubCli  = Send<i32, Send<i32, Recv<i32, Eps>>>;
+        type CalcCli = Choose<AddCli, SubCli>;
+
+        fn client(c: Chan<(), CalcCli>) {
+            let c = c.sel1();
+            let c = c.send(42);
+            let c = c.send(1);
+            let (c, n) = c.recv();
+            println!("{}", n);
+            c.close();
+        }
+    }
+
+
     // Calculator server
 
     type CalcSrv =

### 2019-06-28 tests/tests.rs

@@ -31,6 +31,25 @@ fn ping_works() {
 }
 
 
+/// Test writing a program which duplicates a session.
+///
+/// ```compile_fail
+/// assert!(|| -> Result<(), Box<Error>> {
+///
+///     let r1 = fork(move |s1: Send<(), End>| {
+///         let s2 = send((), s1);
+///         close(s2)?;
+///         let s3 = send((), s1);
+///         close(s3)?;
+///         Ok(())
+///     });
+///     let ((), r2) = recv(r1)?;
+///     close(r2)?;
+///     Ok(())
+///
+/// }().is_ok());
+/// ```
+
 // Test a simple calculator server, implemented using binary choice.
 
 type NegServer<N> = Recv<N, Send<N, End>>;
@@ -313,6 +332,7 @@ fn recursion_works() {
     assert!(other_thread.join().is_ok());
 }
 
+
 // Test selection.
 
 #[test]
@@ -323,8 +343,8 @@ fn selection_works() {
 
     for i in 0..10 {
         let (other_thread, s) = fork_with_thread_id(move |s: Send<u64, End>| {
-            sleep(Duration::from_millis(i * 10));
-            let s = send(i, s);
+            sleep(Duration::from_millis(i * 100));
+            let s = send(9 - i, s);
             close(s)
         });
         other_threads.push(other_thread);
@@ -332,7 +352,7 @@ fn selection_works() {
     }
 
     assert!(|| -> Result<(), Box<Error>> {
-        let mut current_index = 0;
+        let mut current_index = 9;
         loop {
             if rs.is_empty() {
                 break Ok(());
@@ -340,14 +360,15 @@ fn selection_works() {
             else {
                 let (i, r) = select_mut(&mut rs)?;
                 close(r)?;
-                assert_eq!(current_index, i);
-                current_index += 1;
+                assert_eq!(current_index, i, "Messages were received out of order.");
+                current_index = current_index.overflowing_sub(1).0; // decrement
             }
         }
-    }().is_ok());
+    }().is_ok(), "Main thread crashed.");
 
     for other_thread in other_threads {
-        assert!(other_thread.join().is_ok())
+        let msg = format!("Thread {:?} crashed.", other_thread);
+        assert!(other_thread.join().is_ok(), msg);
     }
 }
 

### 2019-06-29 doc/ice2019/main.tex
Final draft version of ICE paper

@@
 In concurrent programming, processes commonly exchange information by sending messages on shared channels.
-However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.
+However, this communication often does not follow a specified protocol, and if it does, the correctness of the implementation with respect to that protocol is rarely checked.
 
-Session types are a substructural typing discipline, introduced by Honda~\cite{honda1993}, which describe communication protocols as \emph{channel types}, capturing both the type and the order of messages.
+Session types are a substructural typing discipline, introduced by Honda~\cite{honda1993}, which describe communication protocols as types, capturing both the types and the order of messages.
 Session type systems, then, ensure that programs correctly implement specified protocols.
@@
 In this paper, we present Rusty Variation\footnote{\url{https://github.com/wenkokke/rusty-variation/}} (RV), a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure.
-It is these strong guarantees of correctness, together with out account of failure, which sets us apart from previous work~\cite{jespersen2015}.
+It is these strong guarantees of correctness, together with our account of failure, which sets us apart from previous work~\cite{jespersen2015}.
 We assume some familiarity with Rust.
-For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}.\footnote{\url{https://doc.rust-lang.org/nightly/book/}}
+For an introduction to Rust, we refer the reader to \textit{The Rust Programming Language}~\cite{rust2019}.
 
 We base our work on Exceptional GV by Fowler \etal~\cite[EGV]{fowler2019}.
-Exceptional GV is a descendant of Good Variation~\cite[GV]{wadler2012} which adds support for exceptions.
+Exceptional GV is a descendant of Good Variation~\cite[GV]{wadler2012}, which adds support for exceptions and the cancellation of sessions.
+GV is a \textlambda-calculus with operations for forking threads and sending and receiving values.
 Crucially, EGV has an affine type system, and permits the explicit cancellation of sessions.
@@
 Rust, however, rejects exceptions in favour of monadic errors, \ie, wrapping possibly failing results in error types such as \lstinline{Option<T>} and \lstinline{Result<T,E>}.
-Second, The core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee livelock freedom, or---well---termination.
+Second, the core language of EGV does not include general recursion. Rust does. In the presence of non-termination, we cannot guarantee livelock freedom, or---well---termination.
+
+RV is currently implemented on top of the channels provided by the \lstinline{crossbeam_channel} crate, but it can be implemented using any library which offers one-shot channels and selection.
 
@@
 
-\section{Contribution: Rusty Variation}\label{sec:contribution}
+\section{Rusty Variation}\label{sec:contribution}
 
@@
 Following EGV, Rusty Variation also permits the cancellation of sessions.
-However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/nomicon/destructors.html}}.
+However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/1.35.0/book/ch15-03-drop.html}}.
 
@@
 This method can be customised by implementing the \lstinline{Drop} trait.
-For instance, if we use the implementation of \lstinline{Drop} shown on the left, running the code on the right would print ``End Dropped!'' twice, once from the child thread, and once from the main thread.
+For instance, if we use the implementation of \lstinline{Drop} shown on the left, running the code on the right would print ``End dropped!'' twice, once from the child thread, and once from the main thread.
 
@@
   fn drop(&mut self) {
-    println!("End Dropped!")
+    println!("End dropped!")
   }
@@
 The \lstinline{choose_left}, \lstinline{choose_right}, and \lstinline{offer_either} functions let processes offer or make a \emph{binary} choice.
-These are \emph{derived} constructs~\cite{honda1993,dardha2017}---they are implemented using only the other functions in our library.
-They send or receive a value of the sum type, \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.
+These are \emph{derived} constructs~\cite{honda1993,dardha2017}---they are implemented using only the functions in our library---\lstinline{new}, \lstinline{send}, \lstinline{recv}, \lstinline{close}, and \lstinline{cancel}---and the sum type \lstinline{Either}.
+They send or receive a value of the type \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.
 \begin{lstlisting}
@@
 
-\subsection{Selection}\label{sec:selection}
+\subsection{Homogeneous Selection}\label{sec:selection}
 Selection is a mechanism which allows processes to block on a list of actions, waiting until the first of them fires.
 Writing programs which use selection can be tricky and error prone, even using libraries specifically built to help with selection.
-For instance, the following is an excerpt from the documentation of \lstinline{crossbeam_channel::Select}\footnote{\url{https://docs.rs/crossbeam-channel/0.1.2/crossbeam_channel/struct.Select.html}}, a construct which implements a selection mechanism:
+For instance, the following is an excerpt from the documentation of \lstinline{crossbeam_channel::Select}\footnote{\url{https://docs.rs/crossbeam-channel/0.3.8/crossbeam_channel/struct.Select.html}}, a construct which implements a selection mechanism:
 \begin{quotation}
   \noindent
-  Rules which must be respected in order for selection to work properly:
-  \begin{enumerate}
-  \item Before each selection, a fresh Select must be created.
-  \item Selection cases must be repeatedly probed in a loop.
-  \item If a selection case fires, the loop must be broken without probing cases any further.
-  \item In each iteration of the loop, the same set of cases must be probed in the same order.
-  \item No selection case may be repeated.
-  \item No two cases may operate on the same end (receiving or sending) of the same channel.
-  \item There must be at least one send, or at least one recv case.
-  \end{enumerate}
+  Once a list of operations has been built with \lstinline{Select}, there are two different ways of proceeding:
+  \begin{itemize}
+  \item Select an operation with \lstinline{try_select}, \lstinline{select}, or \lstinline{select_timeout}.
+If successful, the returned selected operation has already begun and \emph{must} be completed.
+If we don't complete it, a panic will occur.
+  \item Wait for an operation to become ready with \lstinline{try_ready}, \lstinline{ready}, or \lstinline{ready_timeout}.
+If successful, we may attempt to execute the operation, but are not obliged to.
+In fact, it's possible for another thread to make the operation not ready just before we try executing it, so it's wise to use a retry loop.
+However, note that these methods might return with success spuriously, so it's a good idea to always double check if the operation is really ready.
+  \end{itemize}
 \end{quotation}
-None of these rules are enforced by Rust---instead, all of them must be checked by the programmer---and bugs in implementing selection can be hard to find.
+None of this advice is enforced by Rust---instead, this must be checked by the programmer---and bugs in the use of selection can be hard to find, especially if the approach using \lstinline{ready} is used.
 
@@
 The \lstinline{select_mut} and \lstinline{select} functions are homogeneous: they only select over vectors of endpoints with \emph{the same type}.
-However, this does not rule out heterogeneous uses of selection, where the operations are of different types.
-It is possible to perform heterogeneous selection by defining an enumeration, in the style of the \lstinline{CalcOp} type mentioned in~\cref{sec:choice}.
-
-TODO: THIS IS WHERE I STOPPED
+This rules out heterogeneous uses of selection, where the operations are of different types.
+Selection is not a part of EGV, and hence our strong correctness guarantees do not extend to selection.
 
 \subsection{Implementation}\label{sec:implementation}
-We use the asynchronous channels from \lstinline{crossbeam_channel}.
-These are ideal, as they already implement the correct cancellation behaviour.
+So far we have avoided discussing the implementation whenever possible, and with good reason: RV can be implemented using \emph{any} library which offers one-shot channels and selection.
+Hence, the important contribution in this paper is not the implementation, but the programming interface offered by RV.
+
+Let us briefly discuss the implementation.
+We use the asynchronous channels from \lstinline{crossbeam_channel}, which at the time of writing is the de facto implementation of channels in Rust.
 Channels are typed, and can only transfer values of a single type.
-Therefore, we use one-shot channels, and encode sessions following Scalas and Yoshida, and Padovani~\cite{scalas2016,padovani2017}: each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value, and each \lstinline{recv} receives a value together with a channel on which to continue the session.
-Fortunately, the channels in \lstinline{crossbeam_channel} are highly optimised for one-shot usage.
+Therefore, we use one-shot channels, and encode sessions following Scalas and Yoshida, and Padovani~\cite{scalas2016,padovani2017}: Each \lstinline{send} creates a new session for the continuation, and uses the primitive \lstinline{send} operation to send the value along with one endpoint of the continuation session.
+If the primitive \lstinline{send} operation returns an error, we swallow it, using \lstinline{unwrap_or}.
+Each \lstinline{recv} receives a value together with a channel on which to continue the session.
+If the primitive \lstinline{recv} operation returns an error, the \lstinline{?}-operator short-circuits, and we return the error.
+The channels in \lstinline{crossbeam_channel} are highly optimised for one-shot usage.
+\begin{lstlisting}
+use crossbeam_channel::{Sender, Receiver}; // abbreviated import list
+
+pub struct Send<T, S: Session> { channel: Sender<(T, S::Dual)> }
+pub struct Recv<T, S: Session> { channel: Receiver<(T, S)> }
+
+pub fn send<T, S: Session>(x: T, s: Send<T, S>) -> S
+{
+    let (here, there) = S::new();
+    s.channel.send((x, there)).unwrap_or(());
+    here
+}
+
+pub fn recv<T, S: Session>(s: Recv<T, S>) -> Option<(T, S)>
+{
+    let (v, s) = s.channel.recv()?;
+    Some((v, s))
+}
+\end{lstlisting}
+Closing is implemented using two asynchronous channels, which we use to simulate a single synchronous channel.
+Both channels transmit a unit value.
+Each close operator first sends on one of the channels (for which it has the sender) and then receives (blocking) on the other channel (for which it has the receiver).
+If the other endpoint is cancelled, both channels are marked as disconnected, and thus any close operation which is still blocking will raise an exception.
+Otherwise, the first close operation blocks until the second is executed.
+\begin{lstlisting}
+pub struct End { sender: Sender<()>, receiver: Receiver<()> }
+
+pub fn close(s: End) -> Option<()> {
+    s.sender.send(()).unwrap_or(());
+    s.receiver.recv()?;
+    Some(())
+}
+\end{lstlisting}
+While the current implementation is efficient, it could likely be improved by replacing \lstinline{crossbeam_channel} with a minimal implementation of one-shot channels and selection.
+
+The Rust standard library offers \lstinline{std::sync::mpsc}, an implementation of channels.
+The first implementation of RV used the channels from this library.
+However, \lstinline{std::sync::mpsc} has some fundamental issues with its design\footnote{\url{https://github.com/rust-lang/rust/pull/42397\#issuecomment-315867774}}, and seems to be on its way out.
+In 2018, the selection mechanism, \lstinline{mpsc_select!}, was deprecated\footnote{\url{https://github.com/rust-lang/rust/issues/27800}}, and there are plans to phase out \lstinline{std::sync::mpsc} altogether~\cite{glavina2019}.
 
 \subsection{What If I Want to Be Evil?}\label{sec:limitations}
-There are three ways in which you can make code written using Rusty Variation lock.
-The first is by looping, which we discussed earlier.
-The other two are explicit, and surprisingly, \lstinline{unsafe} is not one of them.
+There are three ways in which you can make code written using Rusty Variation lock in safe Rust. The first is by writing an infinite loop, which we discussed earlier.
 
-The biggest issue lies in the fact that Rust's semantics do not guarantee that destructors are actually run.
-This is important for manually managed memory, \eg doubly linked lists.
-For this reason, it is always possible to ``forget'' a value, using \lstinline{std::mem::forget}, without running its destructors.
+The biggest issue lies in the fact that Rust's semantics do not guarantee that the destructors are actually run.
+This does not mean that Rust can \emph{arbitrarily} decide not to run destructors, but that it is possible to write programs which leak memory.
+The most obvious example of this is \lstinline{std::mem::forget}\footnote{\url{https://doc.rust-lang.org/1.35.0/std/mem/fn.forget.html}}:
+\begin{quotation}
+  \noindent
+  Takes ownership and ``forgets'' about the value \emph{without running its destructor}.
 
-Secondly, it is not possible---at the time of writing---to export a trait without exporting all its members.
-Therefore, any user can use the \lstinline{new} function, and doing so, it is trivial to write a program which deadlocks.
+  \noindent
+  Any resources the value manages, such as heap memory or a file handle, will linger forever in an unreachable state. However, it does not guarantee that pointers to this memory will remain valid.
+\end{quotation}
+Explicitly calling \lstinline{std::mem::forget} on session endpoints is unlikely to happen by accident, hence we don't consider this a huge shortcoming.
+However, it is also possible to leak memory by creating reference cycles\footnote{\url{https://doc.rust-lang.org/1.35.0/book/ch15-06-reference-cycles.html}}---in fact, \lstinline{std::mem::forget} was initially marked as \lstinline{unsafe}, until it was discovered that it could be implemented in safe Rust creating reference cycles.
+This means that care needs to be taken when combining sessions and cyclic data structures, \eg storing session endpoints in doubly linked lists.
 
+The second problem arises from the availability of \lstinline{new}, which creates two dual session endpoints.
+Code written using only \lstinline{fork} is guaranteed to be deadlock free~\cite{lindley2015}, but using \lstinline{new} and \lstinline{std::thread::spawn}, it becomes easy to write code which deadlocks.
+However, it is impossible to export a trait without exporting all its members.
+Arguably, hiding \lstinline{new} is also undesirable.
+Using only \lstinline{fork}, it is impossible to create any cyclic communication structures, including those which are well behaved.
+Therefore, a programmer may occasionally \emph{need} access to \lstinline{new}.
+
+The following two programs illustrate the two ways of constructing deadlocking programs discussed above.
+The program on the left uses \lstinline{std::mem::forget} to drop the session endpoint without running its destructors.
+The program on the right creates two pairs of channels to construct a deadlock:
+\vspace{-0.5\baselineskip}
 \begin{minipage}[t]{0.5\linewidth}
@@
 Laumann \etal~\cite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by Gay and Vasconcelos~\cite{gay2009}.
-Their implementation guarantees that programs correctly implement specified communication protocols with some caveats.
-LAST depends upon linear types, meaning that sessions have to be used exactly once.
-Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped.
+Their implementation guarantees session fidelity, albeit with some caveats.
+LAST is a linear language, meaning that sessions can neither be duplicated nor dropped.
+As we discussed in~\cref{sec:cancellation}, Rust is an affine language, meaning that values---and, particularly relevant here, sessions---can be dropped.
 The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}.
-In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \lstinline{#[must_use]} annotation.
-However, in general, this is not something the compiler can guarantee.
-Furthermore, linear session types ignore the reality that channels may be dropped at any time due to, \eg a thread panicking.
 
-We embrace the affine, and base our work on Exceptional GV by Fowler \etal~\cite[EGV]{fowler2019}.
-Exceptional GV is a descendant of Good Variation~\cite[GV]{wadler2012}, which itself is a descendant of LAST.
-Crucially, EGV has an affine type system, and allows for the explicit cancellation of sessions.
-Furthermore, EGV offers a stronger metatheory than LAST.
-In addition to session fidelity, EGV also guarantees global progress, freedom from deadlocks and livelocks, confluence, and termination.
-That is to say, there are no locks or race conditions, and all programs eventually halt.
-For a discussion of EGV and its metatheory, we refer the reader to the work by Fowler \etal~\cite{fowler2019}.
+Laumann \etal~\cite{jespersen2015} remark that they ``cannot prevent a channel value from being dropped prematurely''.
+They offer several solutions, but remark that each of these solutions are either insufficient, premature, or both.
+They note that the proposal to add linear types to Rust is currently postponed\footnote{\url{https://github.com/rust-lang/rfcs/pull/776}}, and that while \lstinline{#[must_use]} can be made to issue a warning if a value is dropped, it is ``easy to accidentally and purposely bypass''.
+Finally, they propose a compiler extension, developed in collaboration with Manish Goregaokar, \lstinline{humpty_dumpty}\footnote{\url{https://github.com/Manishearth/humpty_dumpty}}, ``which provides a lint that ensures that types annotated  with \lstinline{#[drop_protect]} are used linearly''.
+However, they note that \lstinline{humpty_dumpty} ``is still a work in progress, which by no means guarantees linearity in all cases''.
+No further work has been done on \lstinline{humpty_dumpty} since 2015.
+
+We argue that even if Rust were able to enforce linearity, the \emph{idea} of linear session types is flawed (see~\cref{sec:cancellation}), as sessions may be dropped at any time, for any number of reasons, \eg a process panicking, dropping all sessions in scope.
+
+The following two programs illustrate the difference in handling failure between Rusty Variation and the work by Laumann \etal~\cite{jespersen2015}.
+In both programs, the programmer has forgotten to complete the server, accidentally dropping the session endpoint \lstinline{_s} or channel \lstinline{_c}, respectively.
+The program on the left, written in RV, returns an error.
+The program on the right, written using the library described by Laumann \etal\cite{jespersen2015}, segfaults.
+We have since reported this segfault to Laumann \etal, and they have fixed it in the most recent version of the library.
+The program on the right now panics:
 
@@
 
+Another feature which distinguishes Rusty Variation from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice.
+Rusty Variation uses enumerations to provide labelled choice, as described in~\cref{sec:choice}.
+Laumann \etal~\cite{jespersen2015} provide an \lstinline{offer!} macro, which expands a labelled offer to a series of binary offers, \eg offering a ternary choice expands to two nested invocations of \lstinline{offer_either}.
+
+The \lstinline{offer!} macro allows---even requires---the programmer to label their choices, but during macro expansion these labels are simply discarded \emph{without being checked}.
+This allows us to write the program below, in which an inconsistency between types and documentation on the one hand, and the labelled implementation on the other hand, results in a bug.
+The example program implements a calculator server, which offers the choice between squaring and negation, in that order.
+
+In the \lstinline{server} function, the order in which the \lstinline{NEG}-case and the \lstinline{SQR}-case are handled doesn't match the order in which they are declared in the types.
+However, the code compiles, not because the cases are associated correctly through the use of labels, but because the labels aren't checked, and squaring and negation happen to have the same type signature.
+
+When the client selects the first option, they wrongly assume it represents squaring, based on the types.
+Unfortunately, it represents negation.
+The \lstinline{main} program, when run, therefore execute negation, and erroneously returns \lstinline{-42}:
+
 \vspace{-0.5\baselineskip}
@@
 // Laumann et al.
-// The order in the type doesn't match
-// the order in the code. This compiles,
-// not because the cases are associated
-// correctly, but because the labels are
-// not checked.
-
 type SqrSrv  = Recv<i32, Send<i32, Eps>>;
@@
 
-// The client selects the first option,
-// assuming it represents squaring...
-// Unfortunately, it represents negation.
-
 type SqrCli  = Send<i32, Recv<i32, Eps>>;
@@
 
+// main() prints -42
 fn main() {
-  connect(server, client); // prints -42
+  connect(server, client);
 }
@@
 
-\subsection{Future Work}\label{sec:future-work}
-In \cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.
+Lastly, while we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices.
+Below we show two definitions of the session type for a stream of integers, one in Rusty Variation, and one using the library by Laumann \etal~\cite{jespersen2015}:
+
+\begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
+// Rusty Variation
+type IntStream = Send<i32, IntStream>;
+\end{lstlisting}  
+\end{minipage}%
+\begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
+// Laumann et al.
+type IntStream = Rec<Send<i32, Var<Z>>>;
+\end{lstlisting}  
+\end{minipage}
+
+\noindent
+The usage of de Bruijn indices to encode recursive session types is an artefact of the fact that, when Laumann \etal~\cite{jespersen2015} implemented their library, Rust didn't support recursive types.
+
+\section{Future Work}\label{sec:future-work}
+
+\subsection{Heterogeneous Selection}
+In~\cref{sec:selection}, we mention that the \lstinline{select_mut} and \lstinline{select} functions are homogeneous: they only select over vectors of endpoints with \emph{the same type}.
+Heterogeneous selection is an important operation for channel-based communication.
+\emph{Typing} heterogeneous selection, and by extension, heterogeneous vectors is notoriously cumbersome without dependent types.
+There are several avenues for future research:
+
+The \lstinline{frunk} crate\footnote{\url{https://beachape.com/frunk/frunk/index.html}} provides an encoding of heterogeneous lists in Rust.
+We could use this encoding to define a heterogeneous variant of selection.
+
+Rust has limited support for existential types\footnote{\url{https://github.com/rust-lang/rfcs/blob/master/text/2071-impl-trait-existential-types.md}}, which could be used to quantify over the types of the values received and the session continuations, and select on a homogeneous vector of the (purely fictional) type \lstinline[mathescape=true]{Vec<$\exists$T$.\exists$S$.$Recv<T,S>>}.
+
+\subsection{Testing \& Verification}\label{sec:verification}
+In~\cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.
 
@@
 
+\subsection{Extending Exceptional GV}
+Several features of Rusty Variation are not covered by the formal language EGV.
+In~\cref{sec:selection}, we mention that selection isn't a part of EGV.
+In~\cref{sec:related-work}, we mention that recursion isn't part of EGV.
+If we wish to have any hope of formally verifying RV, along the lines described in~\cref{sec:verification}, we would need to extend EGV to cover these features.
+
+\clearpage
 \bibliographystyle{eptcs}

### 2019-06-29 doc/ice2019/main.tex
Removed emdash

@@
 The \lstinline{choose_left}, \lstinline{choose_right}, and \lstinline{offer_either} functions let processes offer or make a \emph{binary} choice.
-These are \emph{derived} constructs~\cite{honda1993,dardha2017}---they are implemented using only the functions in our library---\lstinline{new}, \lstinline{send}, \lstinline{recv}, \lstinline{close}, and \lstinline{cancel}---and the sum type \lstinline{Either}.
+These are \emph{derived} constructs~\cite{honda1993,dardha2017}.
+They are implemented using only the functions in our library---\lstinline{new}, \lstinline{send}, \lstinline{recv}, \lstinline{close}, and \lstinline{cancel}---and the sum type \lstinline{Either}.
 They send or receive a value of the type \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session.

### 2019-06-29 src/lib.rs
Added dyn; updated README

@@ -82,7 +82,7 @@ where
     here
 }
 
-pub fn recv<T, S>(s: Recv<T, S>) -> Result<(T, S), Box<Error>>
+pub fn recv<T, S>(s: Recv<T, S>) -> Result<(T, S), Box<dyn Error>>
 where
     T: marker::Send,
     S: Session,
@@ -99,7 +99,7 @@ pub fn cancel<T>(x: T) -> () {
     mem::drop(x);
 }
 
-pub fn close(s: End) -> Result<(), Box<Error>> {
+pub fn close(s: End) -> Result<(), Box<dyn Error>> {
     s.sender.send(()).unwrap_or(());
     s.receiver.recv()?;
     Ok(())
@@ -111,7 +111,7 @@ pub fn close(s: End) -> Result<(), Box<Error>> {
 pub fn fork_with_thread_id<S, P>(p: P) -> (JoinHandle<()>, S::Dual)
 where
     S: Session + 'static,
-    P: FnOnce(S) -> Result<(), Box<Error>> + marker::Send + 'static
+    P: FnOnce(S) -> Result<(), Box<dyn Error>> + marker::Send + 'static
 {
     let (there, here) = Session::new();
     let other_thread = spawn(move || {
@@ -129,7 +129,7 @@ where
 pub fn fork<S, P>(p: P) -> S::Dual
 where
     S: Session + 'static,
-    P: FnOnce(S) -> Result<(), Box<Error>> + marker::Send + 'static
+    P: FnOnce(S) -> Result<(), Box<dyn Error>> + marker::Send + 'static
 {
     fork_with_thread_id(p).1
 }
@@ -143,12 +143,12 @@ pub type Choose<S1, S2> =
     Send<Either<<S1 as Session>::Dual, <S2 as Session>::Dual>, End>;
 
 pub fn offer_either<'a, S1, S2, F, G, R>(s: Offer<S1, S2>, f: F, g: G)
-                                         -> Result<R, Box<Error + 'a>>
+                                         -> Result<R, Box<dyn Error + 'a>>
 where
     S1: Session,
     S2: Session,
-    F: FnOnce(S1) -> Result<R, Box<Error + 'a>>,
-    G: FnOnce(S2) -> Result<R, Box<Error + 'a>>,
+    F: FnOnce(S1) -> Result<R, Box<dyn Error + 'a>>,
[…]

### 2019-06-29 tests/tests.rs

@@ -16,7 +16,7 @@ use std::time::Duration;
 
 #[test]
 fn ping_works() {
-    assert!(|| -> Result<(), Box<Error>> {
+    assert!(|| -> Result<(), Box<dyn Error>> {
 
         let s = fork(move |s: Send<(), End>| {
             let s = send((), s);
@@ -34,7 +34,7 @@ fn ping_works() {
 /// Test writing a program which duplicates a session.
 ///
 /// ```compile_fail
-/// assert!(|| -> Result<(), Box<Error>> {
+/// assert!(|| -> Result<(), Box<dyn Error>> {
 ///
 ///     let r1 = fork(move |s1: Send<(), End>| {
 ///         let s2 = send((), s1);
@@ -61,7 +61,7 @@ type AddClient<N> = <AddServer<N> as Session>::Dual;
 type SimpleCalcServer<N> = Offer<NegServer<N>, AddServer<N>>;
 type SimpleCalcClient<N> = <SimpleCalcServer<N> as Session>::Dual;
 
-fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<Error>> {
+fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<dyn Error>> {
     offer_either(s,
                  |s: NegServer<i32>| {
                      let (x, s) = recv(s)?;
@@ -80,7 +80,7 @@ fn simple_calc_server(s: SimpleCalcServer<i32>) -> Result<(), Box<Error>> {
 
 #[test]
 fn simple_calc_works() {
-    assert!(|| -> Result<(), Box<Error>> {
+    assert!(|| -> Result<(), Box<dyn Error>> {
 
         let mut rng = thread_rng();
 
@@ -123,7 +123,7 @@ enum CalcOp<N: marker::Send> {
 type NiceCalcServer<N> = Recv<CalcOp<N>, End>;
 type NiceCalcClient<N> = <NiceCalcServer<N> as Session>::Dual;
 
-fn nice_calc_server(s: NiceCalcServer<i32>) -> Result<(), Box<Error>> {
+fn nice_calc_server(s: NiceCalcServer<i32>) -> Result<(), Box<dyn Error>> {
     offer!(s, {
         CalcOp::Neg(s) => {
             let (x, s) = recv(s)?;
@@ -143,7 +143,7 @@ fn nice_calc_server(s: NiceCalcServer<i32>) -> Result<(), Box<Error>> {
 
 #[test]
 fn nice_calc_works() {
[…]

### 2019-06-29 tests/benchmark_tests.rs
Updated benchmark tests

@@ -27,7 +27,7 @@ mod rusty_variation_bench {
     type CalcSrv = Recv<CalcOp, End>;
     type CalcCli = <CalcSrv as Session>::Dual;
 
-    fn calc_server(s: CalcSrv) -> Result<(), Box<Error>> {
+    fn calc_server(s: CalcSrv) -> Result<(), Box<dyn Error>> {
         ::rusty_variation::offer!(s, {
             CalcOp::CLOSE(s) => {
                 close(s)?;
@@ -58,7 +58,7 @@ mod rusty_variation_bench {
         })
     }
 
-    fn neg_client(s: CalcCli) -> Result<(), Box<Error>> {
+    fn neg_client(s: CalcCli) -> Result<(), Box<dyn Error>> {
         let n = thread_rng().gen();
         let s = ::rusty_variation::choose!(CalcOp::NEGATE, s);
         let s = send(n, s);
@@ -84,46 +84,6 @@ mod session_types_bench {
     use rand::{Rng,thread_rng};
     use test::Bencher;
     use session_types::*;
-
-    mod secret {
-
-        use session_types::*;
-
-        type AddSrv  = Recv<i32, Recv<i32, Send<i32, Eps>>>;
-        type SubSrv  = Recv<i32, Recv<i32, Send<i32, Eps>>>;
-        type CalcSrv = Offer<AddSrv, SubSrv>;
-
-        fn server(c: Chan<(), CalcSrv>) {
-            offer!{c,
-                   SUB => {
-                       let (c, n) = c.recv();
-                       let (c, m) = c.recv();
-                       let c = c.send(n - m);
-                       c.close();
-                   },
-                   ADD => {
-                       let (c, n) = c.recv();
-                       let (c, m) = c.recv();
-                       let c = c.send(n + m);
-                       c.close();
-                   }
-            }
-        }
-
-        type AddCli  = Send<i32, Send<i32, Recv<i32, Eps>>>;
-        type SubCli  = Send<i32, Send<i32, Recv<i32, Eps>>>;
-        type CalcCli = Choose<AddCli, SubCli>;
-
-        fn client(c: Chan<(), CalcCli>) {
-            let c = c.sel1();
-            let c = c.send(42);
-            let c = c.send(1);
-            let (c, n) = c.recv();
-            println!("{}", n);
-            c.close();
-        }
-    }
-
 
     // Calculator server
 

### 2019-06-29 doc/ice2019/main.tex
Added a few lines about destructor bombs

@@
 No further work has been done on \lstinline{humpty_dumpty} since 2015.
+The current version of the library enforces linearity \emph{dynamically}, using ``destructor bombs'', implemented by Manish Goregaokar\footnote{\url{https://github.com/Munksgaard/session-types/commit/0f25ccb7c3bc9f65fa8eaf538233e8fe344a189a}}.
+This technique hides a \lstinline{panic!} in channel destructors.
+Each invocation of \lstinline{close} then calls \lstinline{std::mem::forget}, bypassing the destructor bomb.
 

### 2019-06-29 doc/ice2019/main.tex
Minor layout changes

@@
 \end{minipage}
+\vspace{-0.25\baselineskip}
 
@@
 \end{lstlisting}
-\end{minipage}%
+\end{minipage}
+\vspace{-0.25\baselineskip}
 
@@
 \end{lstlisting}
-\end{minipage}
+\end{minipage}%
 \begin{minipage}[t]{0.5\linewidth}
@@
 The current version of the library enforces linearity \emph{dynamically}, using ``destructor bombs'', implemented by Manish Goregaokar\footnote{\url{https://github.com/Munksgaard/session-types/commit/0f25ccb7c3bc9f65fa8eaf538233e8fe344a189a}}.
-This technique hides a \lstinline{panic!} in channel destructors.
+This technique hides a \lstinline{panic!} in channel destructors, which crashes the program if the destructor ever runs.
 Each invocation of \lstinline{close} then calls \lstinline{std::mem::forget}, bypassing the destructor bomb.
@@
 \end{lstlisting}
+\end{minipage}
+\vspace{-0.25\baselineskip}
+
+While we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices.
+Below we show two definitions of the session type for a stream of integers, one in Rusty Variation, and one using the library by Laumann \etal~\cite{jespersen2015}:
+
+\begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
+// Rusty Variation
+type IntStream = Send<i32, IntStream>;
+\end{lstlisting}  
 \end{minipage}%
+\begin{minipage}{0.5\linewidth}
+\begin{lstlisting}
+// Laumann et al.
+type IntStream = Rec<Send<i32, Var<Z>>>;
+\end{lstlisting}  
+\end{minipage}
+\vspace{-0.25\baselineskip}
 
-Another feature which distinguishes Rusty Variation from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice.
+\noindent
+The usage of de Bruijn indices to encode recursive session types is an artefact of the fact that, when Laumann \etal~\cite{jespersen2015} implemented their library, Rust didn't support recursive types.
+
+A last feature which distinguishes Rusty Variation from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice.
 Rusty Variation uses enumerations to provide labelled choice, as described in~\cref{sec:choice}.
@@
 type CalcSrv = Offer<SqrSrv, NegSrv>;
+\end{lstlisting}
+\end{minipage}%
+\begin{minipage}[t]{0.5\linewidth}
+\begin{lstlisting}
 
+type SqrCli  = Send<i32, Recv<i32, Eps>>;
+type NegCli  = Send<i32, Recv<i32, Eps>>;
+type CalcCli = Choose<SqrCli, NegCli>;
+\end{lstlisting}
+\end{minipage}
+\vspace{-0.25\baselineskip}
+
+\vspace{-0.5\baselineskip}
+\begin{minipage}[t]{0.5\linewidth}
+\begin{lstlisting}
 fn server(c: Chan<(), CalcSrv>) {
@@
 \begin{lstlisting}
-
-type SqrCli  = Send<i32, Recv<i32, Eps>>;
-type NegCli  = Send<i32, Recv<i32, Eps>>;
-type CalcCli = Choose<SqrCli, NegCli>;
-
 fn client(c: Chan<(), CalcCli>) {
@@
 \end{minipage}
-
-Lastly, while we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices.
-Below we show two definitions of the session type for a stream of integers, one in Rusty Variation, and one using the library by Laumann \etal~\cite{jespersen2015}:
-
-\begin{minipage}{0.5\linewidth}
-\begin{lstlisting}
-// Rusty Variation
-type IntStream = Send<i32, IntStream>;
-\end{lstlisting}  
-\end{minipage}%
-\begin{minipage}{0.5\linewidth}
-\begin{lstlisting}
-// Laumann et al.
-type IntStream = Rec<Send<i32, Var<Z>>>;
-\end{lstlisting}  
-\end{minipage}
-
-\noindent
-The usage of de Bruijn indices to encode recursive session types is an artefact of the fact that, when Laumann \etal~\cite{jespersen2015} implemented their library, Rust didn't support recursive types.
+\vspace{-0.25\baselineskip}
 
@@
 
-\clearpage
 \bibliographystyle{eptcs}

### 2019-11-20 doc/ice2019/main.tex
Rebranded, deleted some slides.

@@
 
-We argue that even if Rust were able to enforce linearity, the \emph{idea} of linear session types is flawed (see~\cref{sec:cancellation}), as sessions may be dropped at any time, for any number of reasons, \eg a process panicking, dropping all sessions in scope.
+We argue that even if Rust were able to enforce linearity statically, the \emph{idea} of linear session types is flawed (see~\cref{sec:cancellation}), as sessions may be dropped at any time, for any number of reasons, \eg a process panicking, dropping all sessions in scope.
 
@@
 In both programs, the programmer has forgotten to complete the server, accidentally dropping the session endpoint \lstinline{_s} or channel \lstinline{_c}, respectively.
-The program on the left, written in RV, returns an error.
+The program on the left, written in RV, returns \lstinline{None}.
 The program on the right, written using the library described by Laumann \etal\cite{jespersen2015}, segfaults.

### 2019-11-20 src/lib.rs

@@ -1,3 +1,5 @@
+#![feature(never_type)]
+
 extern crate crossbeam_channel;
 extern crate either;
 
@@ -206,7 +208,7 @@ macro_rules! choose {
 }
 
 
-// Selection
+// Homogeneous Selection
 
 #[derive(Debug)]
 enum SelectError {

### 2019-11-20 tests/benchmark_tests.rs

@@ -1,18 +1,18 @@
 #![feature(test)]
 
-extern crate rusty_variation;
+extern crate sesh;
 extern crate session_types;
 extern crate rand;
 extern crate test;
 
 #[cfg(test)]
-mod rusty_variation_bench {
+mod sesh_bench {
 
     use std::boxed::Box;
     use std::error::Error;
     use rand::{Rng,thread_rng};
     use test::Bencher;
-    use rusty_variation::*;
+    use sesh::*;
 
     // Calculator server
 
@@ -28,7 +28,7 @@ mod rusty_variation_bench {
     type CalcCli = <CalcSrv as Session>::Dual;
 
     fn calc_server(s: CalcSrv) -> Result<(), Box<dyn Error>> {
-        ::rusty_variation::offer!(s, {
+        ::sesh::offer!(s, {
             CalcOp::CLOSE(s) => {
                 close(s)?;
                 Ok(())
@@ -60,11 +60,11 @@ mod rusty_variation_bench {
 
     fn neg_client(s: CalcCli) -> Result<(), Box<dyn Error>> {
         let n = thread_rng().gen();
-        let s = ::rusty_variation::choose!(CalcOp::NEGATE, s);
+        let s = ::sesh::choose!(CalcOp::NEGATE, s);
         let s = send(n, s);
         let (n_, s) = recv(s)?;
         assert_eq!(-n, n_);
-        let s = ::rusty_variation::choose!(CalcOp::CLOSE, s);
+        let s = ::sesh::choose!(CalcOp::CLOSE, s);
         close(s)?;
         Ok(())
     }

### 2019-11-20 tests/tests.rs

@@ -1,8 +1,8 @@
 extern crate rand;
-extern crate rusty_variation;
+extern crate sesh;
 
 use rand::{Rng, thread_rng};
-use rusty_variation::*;
+use sesh::*;
 use std::boxed::Box;
 use std::error::Error;
 use std::marker;

### 2019-11-20 bench/tests/benchmark_tests.rs
Moved benchmarks with session_types to separate package.

@@ -1,0 +1,155 @@
+#![feature(test)]
+
+extern crate sesh;
+extern crate session_types;
+extern crate rand;
+extern crate test;
+
+#[cfg(test)]
+mod sesh_bench {
+
+    use std::boxed::Box;
+    use std::error::Error;
+    use rand::{Rng,thread_rng};
+    use test::Bencher;
+    use sesh::*;
+
+    // Calculator server
+
+    #[allow(dead_code)]
+    enum CalcOp {
+        CLOSE(End),
+        ADD(Recv<i64, Recv<i64, Send<i64, CalcSrv>>>),
+        NEGATE(Recv<i64, Send<i64, CalcSrv>>),
+        SQRT(Recv<f64, Send<f64, CalcSrv>>),
+        EVAL(Recv<fn(i64) -> bool, Recv<i64, Send<bool, CalcSrv>>>)
+    }
+    type CalcSrv = Recv<CalcOp, End>;
+    type CalcCli = <CalcSrv as Session>::Dual;
+
+    fn calc_server(s: CalcSrv) -> Result<(), Box<dyn Error>> {
+        ::sesh::offer!(s, {
+            CalcOp::CLOSE(s) => {
+                close(s)?;
+                Ok(())
+            },
+            CalcOp::ADD(s) => {
+                let (x, s) = recv(s)?;
+                let (y, s) = recv(s)?;
+                let s = send(x + y, s);
+                calc_server(s)
+            },
+            CalcOp::NEGATE(s) => {
+                let (x, s) = recv(s)?;
+                let s = send(-x, s);
+                calc_server(s)
+            },
+            CalcOp::SQRT(s) => {
+                let (x, s) = recv(s)?;
+                let s = send(x.sqrt(), s);
+                calc_server(s)
+            },
+            CalcOp::EVAL(s) => {
+                let (f, s) = recv(s)?;
+                let (x, s) = recv(s)?;
+                let s = send(f(x), s);
+                calc_server(s)
+            },
+        })
+    }
+
+    fn neg_client(s: CalcCli) -> Result<(), Box<dyn Error>> {
+        let n = thread_rng().gen();
+        let s = ::sesh::choose!(CalcOp::NEGATE, s);
+        let s = send(n, s);
+        let (n_, s) = recv(s)?;
+        assert_eq!(-n, n_);
+        let s = ::sesh::choose!(CalcOp::CLOSE, s);
+        close(s)?;
+        Ok(())
[…]

### 2019-11-20 tests/benchmark_tests.rs

@@ -1,155 +1,0 @@
-#![feature(test)]
-
-extern crate sesh;
-extern crate session_types;
-extern crate rand;
-extern crate test;
-
-#[cfg(test)]
-mod sesh_bench {
-
-    use std::boxed::Box;
-    use std::error::Error;
-    use rand::{Rng,thread_rng};
-    use test::Bencher;
-    use sesh::*;
-
-    // Calculator server
-
-    #[allow(dead_code)]
-    enum CalcOp {
-        CLOSE(End),
-        ADD(Recv<i64, Recv<i64, Send<i64, CalcSrv>>>),
-        NEGATE(Recv<i64, Send<i64, CalcSrv>>),
-        SQRT(Recv<f64, Send<f64, CalcSrv>>),
-        EVAL(Recv<fn(i64) -> bool, Recv<i64, Send<bool, CalcSrv>>>)
-    }
-    type CalcSrv = Recv<CalcOp, End>;
-    type CalcCli = <CalcSrv as Session>::Dual;
-
-    fn calc_server(s: CalcSrv) -> Result<(), Box<dyn Error>> {
-        ::sesh::offer!(s, {
-            CalcOp::CLOSE(s) => {
-                close(s)?;
-                Ok(())
-            },
-            CalcOp::ADD(s) => {
-                let (x, s) = recv(s)?;
-                let (y, s) = recv(s)?;
-                let s = send(x + y, s);
-                calc_server(s)
-            },
-            CalcOp::NEGATE(s) => {
-                let (x, s) = recv(s)?;
-                let s = send(-x, s);
-                calc_server(s)
-            },
-            CalcOp::SQRT(s) => {
-                let (x, s) = recv(s)?;
-                let s = send(x.sqrt(), s);
-                calc_server(s)
-            },
-            CalcOp::EVAL(s) => {
-                let (f, s) = recv(s)?;
-                let (x, s) = recv(s)?;
-                let s = send(f(x), s);
-                calc_server(s)
-            },
-        })
-    }
-
-    fn neg_client(s: CalcCli) -> Result<(), Box<dyn Error>> {
-        let n = thread_rng().gen();
-        let s = ::sesh::choose!(CalcOp::NEGATE, s);
-        let s = send(n, s);
-        let (n_, s) = recv(s)?;
-        assert_eq!(-n, n_);
-        let s = ::sesh::choose!(CalcOp::CLOSE, s);
-        close(s)?;
-        Ok(())
[…]

### 2019-11-20 doc/ice2019/main.tex
Replaced occurances of Rusty Variation with Sesh

@@
 
-\title{Rusty Variation\\ Deadlock-free Sessions with Failure in Rust}
+\title{Sesh: Deadlock-free Sessions with Failure in Rust}
 \author{Wen Kokke
@@
 }
-\def\titlerunning{Rusty Variation}
+\def\titlerunning{Sesh}
 \def\authorrunning{Wen Kokke}
@@
 \begin{abstract}
-  Rusty Variation (RV) is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees.
-Programs written using RV are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
+  Sesh is a library for session-typed communication in Rust which offers strong compile-time correctness guarantees.
+Programs written using Sesh are guaranteed to respect a specified protocol, and are guaranteed to be free from deadlocks and races.
 \end{abstract}
@@
 
-In this paper, we present Rusty Variation\footnote{\url{https://github.com/wenkokke/rusty-variation/}} (RV), a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure.
+In this paper, we present Sesh\footnote{\url{https://github.com/wenkokke/sesh/}}, a library for session-typed communication in Rust which allows users to specify and check the adherence to communication protocols, in addition to offering strong correctness guarantees, such as freedom from deadlocks and unwanted races, while accounting for the possibility of failure.
 It is these strong guarantees of correctness, together with our account of failure, which sets us apart from previous work~\cite{jespersen2015}.
@@
 
-Rusty Variation is an implementation of Exceptional GV (EGV) in Rust.
-We claim that RV is a faithful implementation of EGV, preserving much of EGV's metatheory, meaning RV guarantees session fidelity, global progress, confluence, and freedom from deadlocks and unwanted races.
+Sesh is an implementation of Exceptional GV (EGV) in Rust.
+We claim that Sesh is a faithful implementation of EGV, preserving much of EGV's metatheory, meaning Sesh guarantees session fidelity, global progress, confluence, and freedom from deadlocks and unwanted races.
 However, we depart from EGV in two ways:
@@
 
-RV is currently implemented on top of the channels provided by the \lstinline{crossbeam_channel} crate, but it can be implemented using any library which offers one-shot channels and selection.
+Sesh is currently implemented on top of the channels provided by the \lstinline{crossbeam_channel} crate, but it can be implemented using any library which offers one-shot channels and selection.
 
 The paper proceeds as follows.
-In~\cref{sec:contribution}, we describe the programming interface offered by Rusty Variation, and provide several example programs.
-In~\cref{sec:related-work}, we compare Rusty Variation to the related work by Laumann \etal~\cite{jespersen2015}.
-Finally, in~\cref{sec:future-work}, we describe future work on verifying the correctness guarantees offered by RV.
+In~\cref{sec:contribution}, we describe the programming interface offered by Sesh, and provide several example programs.
+In~\cref{sec:related-work}, we compare Sesh to the related work by Laumann \etal~\cite{jespersen2015}.
+Finally, in~\cref{sec:future-work}, we describe future work on verifying the correctness guarantees offered by Sesh.
 
-\section{Rusty Variation}\label{sec:contribution}
+\section{Sesh}\label{sec:contribution}
 
@@
 In~\cref{sec:introduction}, we mentioned that EGV has an affine type system, and permits the explicit cancellation of sessions.
-Following EGV, Rusty Variation also permits the cancellation of sessions.
+Following EGV, Sesh also permits the cancellation of sessions.
 However, instead of having cancellation be an explicit primitive, we implement it implicitly for any dropped value, using Rust's destructors\footnote{\url{https://doc.rust-lang.org/1.35.0/book/ch15-03-drop.html}}.
@@
 
-RV offers an alternative, the \lstinline{select_mut} function.
+Sesh offers an alternative, the \lstinline{select_mut} function.
 This function operates on a vector of \emph{receiving} endpoints, returning the result of the first receive operation which completes, and removing the corresponding session from the vector.
-We restrict ourselves to the receiving endpoints, as RV's \lstinline{send} operation is non-blocking, hence selecting on sending endpoints doesn't make any sense.
+We restrict ourselves to the receiving endpoints, as Sesh's \lstinline{send} operation is non-blocking, hence selecting on sending endpoints doesn't make any sense.
 \begin{lstlisting}
@@
 The \lstinline{select_mut} function is implemented on top of \lstinline{crossbeam_channel::Select}, in a way which respects the rules listed above.
-In addition, RV offers an immutable variant of \lstinline{select_mut}, named \lstinline{select}, which returns the remainder of the endpoints along the with result of the selection.
+In addition, Sesh offers an immutable variant of \lstinline{select_mut}, named \lstinline{select}, which returns the remainder of the endpoints along the with result of the selection.
 Using \lstinline{select_mut}, we can write the following program, which forks off ten threads, each sending an index between 1 and 10, and then repeatedly selects on the receiving endpoints, receiving and printing an index, printing the numbers 1 to 10 in a random order:
@@
 \subsection{Implementation}\label{sec:implementation}
-So far we have avoided discussing the implementation whenever possible, and with good reason: RV can be implemented using \emph{any} library which offers one-shot channels and selection.
-Hence, the important contribution in this paper is not the implementation, but the programming interface offered by RV.
+So far we have avoided discussing the implementation whenever possible, and with good reason: Sesh can be implemented using \emph{any} library which offers one-shot channels and selection.
+Hence, the important contribution in this paper is not the implementation, but the programming interface offered by Sesh.
 
@@
 The Rust standard library offers \lstinline{std::sync::mpsc}, an implementation of channels.
-The first implementation of RV used the channels from this library.
+The first implementation of Sesh used the channels from this library.
 However, \lstinline{std::sync::mpsc} has some fundamental issues with its design\footnote{\url{https://github.com/rust-lang/rust/pull/42397\#issuecomment-315867774}}, and seems to be on its way out.
@@
 \subsection{What If I Want to Be Evil?}\label{sec:limitations}
-There are three ways in which you can make code written using Rusty Variation lock in safe Rust. The first is by writing an infinite loop, which we discussed earlier.
+There are three ways in which you can make code written using Sesh lock in safe Rust. The first is by writing an infinite loop, which we discussed earlier.
 
@@
 
-The following two programs illustrate the difference in handling failure between Rusty Variation and the work by Laumann \etal~\cite{jespersen2015}.
+The following two programs illustrate the difference in handling failure between Sesh and the work by Laumann \etal~\cite{jespersen2015}.
 In both programs, the programmer has forgotten to complete the server, accidentally dropping the session endpoint \lstinline{_s} or channel \lstinline{_c}, respectively.
-The program on the left, written in RV, returns \lstinline{None}.
+The program on the left, written in Sesh, returns \lstinline{None}.
 The program on the right, written using the library described by Laumann \etal\cite{jespersen2015}, segfaults.
@@
 \begin{lstlisting}
-// Rusty Variation
+// Sesh
 // main() returns None
@@
 While we haven't discussed recursive session types in this paper, Laumann \etal~\cite{jespersen2015} use a rather cumbersome encoding for recursive session types, using type-level de Bruijn indices.
-Below we show two definitions of the session type for a stream of integers, one in Rusty Variation, and one using the library by Laumann \etal~\cite{jespersen2015}:
+Below we show two definitions of the session type for a stream of integers, one in Sesh, and one using the library by Laumann \etal~\cite{jespersen2015}:
 
@@
 \begin{lstlisting}
-// Rusty Variation
+// Sesh
 type IntStream = Send<i32, IntStream>;
@@
 
-A last feature which distinguishes Rusty Variation from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice.
-Rusty Variation uses enumerations to provide labelled choice, as described in~\cref{sec:choice}.
+A last feature which distinguishes Sesh from the work by Laumann \etal~\cite{jespersen2015} is in how both libraries generalise binary choice to labelled choice.
+Sesh uses enumerations to provide labelled choice, as described in~\cref{sec:choice}.
 Laumann \etal~\cite{jespersen2015} provide an \lstinline{offer!} macro, which expands a labelled offer to a series of binary offers, \eg offering a ternary choice expands to two nested invocations of \lstinline{offer_either}.
@@
 \subsection{Testing \& Verification}\label{sec:verification}
-In~\cref{sec:contribution}, we claim that RV preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of RV is out of the scope of this paper, it would be remiss of us not substantiate this claim.
+In~\cref{sec:contribution}, we claim that Sesh preserves the metatheory of EGV, and while a full discussion of the testing and formal verification efforts we have undertaken to ensure the correctness of Sesh is out of the scope of this paper, it would be remiss of us not substantiate this claim.
 
-Firstly, RV is unit tested.
+Firstly, Sesh is unit tested.
 We test various use-cases from EGV, from basic operations and their interactions with cancellation, to delegation and recursive session types.
@@
 However, unit tests are error prone, and it is hard (if not impossible) to obtain a set of tests which cover all use-cases of a library.
-Unfortunately, tools such as RustBelt~\cite{jung2017} and Rust Distilled~\cite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of RV.
+Unfortunately, tools such as RustBelt~\cite{jung2017} and Rust Distilled~\cite{weiss2018} are not yet mature enough to allow us to \emph{prove} the correctness of Sesh.
 
-In the absence of a mature formal semantics for Rust, we are designing a formal language, RV (or ``formal RV''), which matches, as close as possible, the semantics of our Rust library.
+In the absence of a mature formal semantics for Rust, we are designing a formal language, Sesh (or ``formal Sesh''), which matches, as close as possible, the semantics of our Rust library.
 It is a concurrent lambda calculus with heap-based shared-memory semantics.
-We plan to prove that formal RV guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom.
-Furthermore, we plan to show that there exists a translation from EGV to formal RV, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\cite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV.
-This would show that formal RV is a faithful implementation of EGV.
+We plan to prove that formal Sesh guarantees preservation, progress, termination, confluence and the diamond property, and deadlock and livelock freedom.
+Furthermore, we plan to show that there exists a translation from EGV to formal Sesh, based on the monadic reflection of exceptions into the \lstinline{Option} monad~\cite{filinski1994} and the translation of channel-based communication to shared memory, which is in operational correspondence with EGV.
+This would show that formal Sesh is a faithful implementation of EGV.
 
-Lastly, we plan to ensure the correspondence between formal RV and the implementation of RV, using property-based testing.
-We plan to generate a series of random formal RV terms, using Neat~\cite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal RV.
+Lastly, we plan to ensure the correspondence between formal Sesh and the implementation of Sesh, using property-based testing.
+We plan to generate a series of random formal Sesh terms, using Neat~\cite{claessen2015}, and check if their behaviour, when run as Rust programs, corresponds to the behaviour of the term in formal Sesh.
 
 \subsection{Extending Exceptional GV}
-Several features of Rusty Variation are not covered by the formal language EGV.
+Several features of Sesh are not covered by the formal language EGV.
 In~\cref{sec:selection}, we mention that selection isn't a part of EGV.
 In~\cref{sec:related-work}, we mention that recursion isn't part of EGV.
-If we wish to have any hope of formally verifying RV, along the lines described in~\cref{sec:verification}, we would need to extend EGV to cover these features.
+If we wish to have any hope of formally verifying Sesh, along the lines described in~\cref{sec:verification}, we would need to extend EGV to cover these features.
 

### 2019-11-20 src/lib.rs
Did some extremely basic documentation.

@@ -1,3 +1,67 @@
+//! # Sesh 🥂
+//!
+//! Sesh is an implementation of *deadlock-free binary session types with
+//! failure* in Rust.
+//!
+//! What does that mean?
+//!
+//! - The *session types* bit means that it uses Rust types to describe
+//! communication protocols, and ensure that they're implemented correctly.
+//! - The *binary* bit means that for any protocol, there's two parties involved, no
+//! more, no less.
+//! - The *deadlock-free* bit means, well, that if you write your session typed
+//! programs using Sesh, and you don't cheat by using other concurrency
+//! constructs, your can be sure your programs won't deadlock!
+//! - The *with failure* bit means Sesh is aware that sometimes programs may
+//! fail, and that it takes failure into account. If one of the parties involved
+//! in a protocol crashes, like if the thread panics, the Sesh lets the other
+//! party know, to make sure it doesn't end up just waiting forever.
+//!
+//! The author of this package is an academic, so she is contractually obliged
+//! to only come up with uninspiring examples. With that in mind, let's pretend
+//! you want a server that does addition. You send it two numbers, and it sends
+//! you a number back. There's two session types associated with that protocol.
+//! One for the client and one for the server. The type for the server, using
+//! Sesh, would be:
+//!
+//! ```
+//! # extern crate sesh;
+//! # use sesh::*;
+//! #
+//! type AddServer = Recv<i64, Recv<i64, Send<i64, End>>>;
+//! ```
+//!
+//! Session types are always *dual*. If the client sends a number, the server
+//! should be ready to receive a number, otherwise we'd be in trouble.
+//! We can get the session type for the client using duality. This just replaces
+//! all sends by receives, and vice versa:
+//!
+//! ```
+//! # extern crate sesh;
+//! # use sesh::*;
+//! #
+//! # type AddServer = Recv<i64, Recv<i64, Send<i64, End>>>;
+//! type AddClient = <AddServer as Session>::Dual;
+//! ```
+//!
+//! Once you've written down the protocol, the hard part is out of the way.
[…]
